---
title: "A1 - Dataset selection, cleaning, normalization, and differential gene expression"
author: "Christine En-Tse Cheng"
date: "02-10-2026"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs/A1") })
bibliography: A1_files/A1_references.bib    # TODO: insert the citations/bibliography file
link-citations: true            # this is for the citations/bibliography 
nocite: '@*'
---

## 1 Introduction

TODO: introduce the dataset/paper, maybe briefly talk about how/why the dataset was chosen

This report describes the selection, data preparation, normalization, and differential expression analysis of a publicly available human gene expression dataset.

The selected dataset is [GSE166605](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE166605) from the Gene Expression Omnibus (GEO). This RNA-seq dataset investigates TODO: [transcriptional responses of human CD14+ monocytes under experimentally defined conditions involving differential treatment exposure]. The dataset was chosen because TODO

### 1.1 Dataset Selection {#q1}

brief description

A detailed account of the dataset discovery process is documented separately [here](https://github.com/bcb420-2026/Christine_Cheng/wiki/Entry-4-%E2%80%90-Dataset-Selection-(2.3)).

### 1.2 Dataset Background

-   paper background

####  {#q2}

-   control and test conditions of dataset

## 2 Data Exploration

The following sections describes initial steps in the analysis, from accessing the dataset to cleaning the data and performing preliminary exploratory analysis.

First, necessary libraries were installed (if necessary) and loaded into the R session.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      knitr.kable.max_rows = 6)

# install/attach packages
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

if (!require("GEOquery", quietly = TRUE))
  BiocManager::install("GEOquery")

if (!require("readr", quietly = TRUE))
  install.packages("readr")

if (!require("dplyr", quietly = TRUE))
  install.packages("dplyr")

if (!require("biomaRt", quietly = TRUE))
  install.packages("biomaRt")

library(dplyr)
library(biomaRt)

# library(SummarizedExperiment) # TODO: if using data from getRNASeqData
library(readr) # TODO: if reading raw counts file from getGEOSuppFiles
```

### 2.1 Downloading the dataset

The dataset was accessed and downloaded using `GEOquery` and the corresponding GEO series ID. The raw count data file was read directly in R without manual file management, to ensure reproducibility of results.

TODO: Below is a snapshot of the raw matrix: 

```{r}
GSE_ID <- "GSE166605" # constant for GEO series ID to reduce risk of accessing the wrong dataset
```

```{r}
# dataset_se <- GEOquery::getRNASeqData(GSE_ID)
# dataset_se
# metadata(dataset_se)
# rowData(dataset_se)
```

```{r}
# # insert check for previously downloaded dataset in tempdir() or specified destdir
# dataset_gse <- GEOquery::getGEO(GEO = GSE_ID, 
#                                 destdir = "~/projects/Christine_Cheng/A1_files",
#                                 AnnotGPL = TRUE)
# dataset_gse
```

```{r data}
# access raw files from GEO
dataset_suppFiles <- GEOquery::getGEOSuppFiles(GSE_ID,
                                               makeDirectory = FALSE,
                                               baseDir = "~/projects/Christine_Cheng/A1_files")

counts_file <- file.path("~/projects/Christine_Cheng/A1_files",
                        "GSE166605_D1-5_Neg_C_miR_155_RAW_counts.txt.gz")
stopifnot(file.exists(counts_file)) # prevent errors if counts file retrieval fails

# read in raw count file
raw_counts <- readr::read_tsv(counts_file, show_col_types = FALSE)

# names(raw_counts)[1:min(12, ncol(raw_counts))]  # TODO: remove
raw_counts
```

The raw counts file was additionally processed to remove gene IDs and gene names, such that only raw expression values remained: TODO: display ```counts```
```{r}
# processing count file to only contain expression values (remove GeneID and Gene Name columns)
counts <- raw_counts %>%
  dplyr::select(starts_with("D")) %>% # based on GEO accession page, all sample names start with 'D'
  as.data.frame()

rownames(counts) <- raw_counts$GeneID

counts <- as.matrix(counts) # converted to matrix for compatibility with DE analysis packages
storage.mode(counts) <- "integer"
```

Parsing count data to make metadata for design matrix.

```{r metadata}
# Sample name format: D# CD14 control/treatment_name
samples <- colnames(counts)[1:ncol(counts)]
cell_line  <- unlist(lapply(samples,FUN =
                             function(x){unlist(strsplit(x,split = " "))[2]}))       
treatment <- unlist(lapply(samples,FUN = 
                             function(x){unlist(strsplit(x,split = " "))[3]}))     
meta_df <- data.frame(samples, cell_line, treatment)
```

### 2.2 Exploratory Statistics {#q3}

```{r}
dim(counts)
```

### 2.3 Data Trimming/Cleaning

####  {#q6}

####  {#q7}

```{r}
# # QC, filtering and identifier mapping
# # This chapter performs shared QC steps used by both limma-voom and edgeR and DEseq2.
# 
# obj <- readRDS("data/d1_counts_and_meta.rds")
# counts <- obj$counts
# meta <- obj$meta
# mappings <- obj$mappings
# 
# # Create a DGEList (edgeR data container)
# dge <- edgeR::DGEList(counts = counts)
# 
# # Attach metadata
# # (not required, but helpful)
# dge$samples <- cbind(dge$samples, meta)
# 
# dge$samples
# 
# # filter lowly expressed genes
# design <- model.matrix(~ 0 +  treatment+ cellline, data = meta)
# 
# keep <- edgeR::filterByExpr(dge, design)
# dge_f <- dge[keep, , keep.lib.sizes = FALSE]
# 
# dim(dge)
```

## 3 Data Normalization

```{r}
# # TMM normalization
# dge_f <- edgeR::calcNormFactors(dge_f, method = "TMM")
# dge_f$samples
```

```{r}
# # exploratory MDS plot
# current_colors <- RColorBrewer::brewer.pal(n = length(unique(meta$treatment)), "Set2")
# names(current_colors) <- levels(factor(meta$treatment))
# 
# plotMDS(dge_f,
#   labels = meta$treatment,
#   col = current_colors[meta$treatment],
#   main = "MDS plot (colored by treatment)"
# )
# legend("topright",
#   legend = names(current_colors),
#   col = current_colors,
#   pch = 16,
#   bty = "n"
# )
# 
# # " " " - coloured by cell line
# current_colors_cellline <- RColorBrewer::brewer.pal(n = length(unique(meta$cellline)), "Set1")
# names(current_colors_cellline) <- levels(factor(meta$cellline))
# 
# plotMDS(dge_f,
#   labels = meta$cellline,
#   col = current_colors_cellline[meta$cellline],
#   main = "MDS plot (colored by cellline)"
# )
# legend("topright",
#   legend = names(current_colors_cellline),
#   col = current_colors_cellline,
#   pch = 16,
#   bty = "n"
# )
```

```{r}
# # log-CPM for visualization
# lcpm <- edgeR::cpm(dge_f, log = TRUE, prior.count = 1)
# summary(as.vector(lcpm))
```

```{r}
# saveRDS(list(dge_f = dge_f, 
#              meta = meta, 
#              design = design, 
#              lcpm = lcpm, 
#              mappings = mappings),
#         file = "data/d1_qc_objects.rds")
```

### 3.1 Gene ID mapping

```{r id_mapping}
# strip_ensembl_version <- function(x) sub("\\..*$", "", x)
# 
# kable_head <- function(x, n = 5, caption = NULL) {
#   knitr::kable(utils::head(x, n), caption = caption)
# }
# 
# mapped_counts <- data.frame(ensembl_gene_id = rownames(counts),counts)
# ids<- mapped_counts$ensembl_gene_id
# 
# if (any(grepl("^ENSG", strip_ensembl_version(ids)))) {
#   library(biomaRt)
#   ensembl_ids <- unique(strip_ensembl_version(ids))
#   ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
# 
#   map <- getBM(
#     attributes = c("ensembl_gene_id", "hgnc_symbol"),
#     filters = "ensembl_gene_id",
#     values = ensembl_ids,
#     mart = ensembl
#   )
# 
#   kable_head(map, 10, paste(": mapping preview"))
# 
# counts_mapped <- mapped_counts %>%
#     left_join(map, by = "ensembl_gene_id") %>%
#   dplyr::select( ensembl_gene_id,hgnc_symbol, everything())
# 
# 
#   kable_head(counts_mapped[, 1:min(8, ncol(counts_mapped))], 5, paste(": mapped table preview"))
# }
```

```{r}
# # save intermmediate objects
# saveRDS(list(counts = counts, 
#              meta = meta, 
#              mappings = counts_mapped[,1:2]), 
#         file = "data/d1_counts_and_meta.rds")
```

## 4 Differential Expression Analysis

```{r}
# obj <- readRDS("data/d1_qc_objects.rds")
# dge_f <- obj$dge_f
# meta <- obj$meta
# mappings <- obj$mappings
# 
# design <- obj$design
# 
# # Contrasts
# contr <- makeContrasts(
#   `3CTG_vs_NT` = treatment3CTG - treatmentNT,
#   `20CTG_vs_NT` = treatment20CTG - treatmentNT,
#   `20CTG_vs_3CTG` = treatment20CTG - treatment3CTG,
#   levels = design
# )
# 
# # Estimate dispersion and fit QL model
# edge <- edgeR::estimateDisp(dge_f, design)
# fit <- edgeR::glmQLFit(edge, design)
# 
# plotBCV(edge)
# 
# qlf_3_vs_nt <- edgeR::glmQLFTest(fit, contrast = contr[, "3CTG_vs_NT"])
# qlf_20_vs_nt <- edgeR::glmQLFTest(fit, contrast = contr[, "20CTG_vs_NT"])
# 
# edgeR::topTags(qlf_20_vs_nt)
# 
# qlf_20_vs_3 <- edgeR::glmQLFTest(fit, contrast = contr[, "20CTG_vs_3CTG"])
# 
# edgeR::topTags(qlf_20_vs_3)
# 
# tab_3_vs_nt <- edgeR::topTags(qlf_3_vs_nt, n = Inf)$table %>% rownames_to_column("ensembl")
# tab_20_vs_nt <- edgeR::topTags(qlf_20_vs_nt, n = Inf)$table %>% rownames_to_column("ensembl")
# tab_20_vs_3 <- edgeR::topTags(qlf_20_vs_3, n = Inf)$table %>% rownames_to_column("ensembl")
# 
# head(tab_20_vs_nt)
# 
# volcano_df <- tab_20_vs_nt %>%
#   mutate(sig = FDR < 0.05)
# 
# ggplot(volcano_df, aes(x = logFC, y = -log10(PValue), color = sig)) +
#   geom_point(alpha = 0.6) +
#   theme_minimal() +
#   labs(
#     title = "EdgeR: 20CTG vs NT",
#     x = "log2 fold-change",
#     y = "-log10(p-value)"
#   )
# 
# tab <- topTags(qlf_20_vs_nt, n = Inf)$table %>%
#   rownames_to_column("gene") %>%
#   mutate(
#     neglog10P = -log10(PValue),
#     direction = case_when(
#       FDR < 0.05 & logFC > 0 ~ "Up",
#       FDR < 0.05 & logFC < 0 ~ "Down",
#       TRUE ~ "Not sig"
#     )
#   )
# 
# ggplot(tab, aes(x = logFC, y = neglog10P)) +
#   geom_point(aes(color = direction), alpha = 0.7, size = 1.2) +
#   scale_color_manual(values = c(
#     "Up" = "red",
#     "Down" = "blue",
#     "Not sig" = "grey70"
#   )) +
#   geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "steelblue") +
#   geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "steelblue") +
#   theme_minimal() +
#   labs(
#     title = "edgeR volcano: 20CTG vs NT",
#     x = "log2 fold change (logFC)",
#     y = "-log10(p-value)",
#     color = NULL
#   )
# 
# # Keep only the columns we need and ensure they are character
# map_df <- mappings %>%
#   transmute(
#     ensembl = as.character(ensembl_gene_id),
#     hgnc_symbol = as.character(hgnc_symbol)
#   ) %>%
#   distinct(ensembl, .keep_all = TRUE)
# 
# # Join mapping onto the edgeR results table (tab_20_vs_nt)
# tab_20_vs_nt_annot <- tab_20_vs_nt %>%
#   mutate(ensembl = as.character(ensembl)) %>%
#   left_join(map_df, by = "ensembl") %>%
#   mutate(
#     # Use HGNC when available, otherwise fall back to Ensembl
#     display_name = ifelse(!is.na(hgnc_symbol) & hgnc_symbol != "",
#                           hgnc_symbol,
#                           ensembl)
#   )
# 
# # Quick check: how many mapped?
# mean(!is.na(tab_20_vs_nt_annot$hgnc_symbol))
# 
# head(tab_20_vs_nt_annot[, c("ensembl","hgnc_symbol","display_name","logFC","FDR")])
# 
# top_n <- 50  # change to 25/100 as desired
# 
# top_hits_tbl <- tab_20_vs_nt_annot %>%
#   arrange(FDR, PValue) %>%
#   slice_head(n = top_n)
# 
# # Ensembl IDs for subsetting the matrix
# top_hits_ensembl <- top_hits_tbl$ensembl
# 
# # Row labels we want to show on the heatmap
# top_hits_labels <- top_hits_tbl$display_name
# 
# head(top_hits_tbl[, c("ensembl","display_name","logFC","FDR")])
# 
# # logCPM from counts for visualization
# # logCPM for visualization
# logcpm <- edgeR::cpm(counts, log = TRUE, prior.count = 1)
# 
# # Keep only genes present in the expression matrix
# keep <- top_hits_ensembl %in% rownames(logcpm)
# 
# hm_mat <- logcpm[top_hits_ensembl[keep], , drop = FALSE]
# 
# # Create labels aligned to hm_mat rows
# hm_labels <- top_hits_labels[keep]
# 
# # IMPORTANT: avoid duplicate rownames (HGNC symbols can repeat)
# # Make labels unique while preserving readability
# rownames(hm_mat) <- make.unique(hm_labels)
# 
# # Z-score per gene across samples for pattern visibility
# hm_z <- t(scale(t(hm_mat)))
# 
# dim(hm_z)
# 
# meta$cellline <- as.factor(meta$cellline)
# meta$treatment <- as.factor(meta$treatment)
# 
# # Set3 palettes (qualitative)
# treat_levels <- levels(meta$treatment)
# cell_levels  <- levels(meta$cellline)
# 
# treat_cols <- setNames(
#   brewer.pal(max(3, length(treat_levels)), "Set3")[seq_along(treat_levels)],
#   treat_levels
# )
# 
# cell_cols <- setNames(
#   brewer.pal(max(3, length(cell_levels)), "Set2")[seq_along(cell_levels)],
#   cell_levels
# )
# 
# ha <- HeatmapAnnotation(
#   treatment = meta$treatment,
#   cellline  = meta$cellline,
#   col = list(
#     treatment = treat_cols,
#     cellline  = cell_cols
#   ),
#   annotation_legend_param = list(
#     treatment = list(title = "Treatment"),
#     cellline  = list(title = "Cell line")
#   )
# )
# 
# col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))
# 
# Heatmap(
#   hm_z,
#   col = col_fun,
#   top_annotation = ha,
#   show_row_names = FALSE,
#   show_column_names = FALSE,
#   cluster_rows = TRUE,
#   cluster_columns = TRUE,
#   row_title = paste0("Top ", nrow(hm_z), " genes (edgeR: 20CTG vs NT)"),
#   column_title = "Samples"
# )
# 
# Heatmap(
#   hm_z[1:25,],
#   col = col_fun,
#   top_annotation = ha,
#   show_row_names = TRUE,         # <-- now shows HGNC symbols
#   row_names_gp = grid::gpar(fontsize = 8),
#   show_column_names = FALSE,
#   cluster_rows = TRUE,
#   cluster_columns = TRUE,
#   row_title = paste0("Top ", nrow(hm_z), " genes (edgeR: 20CTG vs NT)"),
#   column_title = "Samples"
# )
# 
# saveRDS(
#   list(
#     meta = meta,
#     counts_mapped_filtered = counts_mapped_filtered,
#     fit = fit,
#     qlf_3_vs_nt = qlf_3_vs_nt,
#     qlf_20_vs_nt = qlf_20_vs_nt,
#     qlf_20_vs_3 = qlf_20_vs_3,
#     tab_3_vs_nt = tab_3_vs_nt,
#     tab_20_vs_nt = tab_20_vs_nt,
#     tab_20_vs_3 = tab_20_vs_3
#   ),
#   file = "data/d1_edger_results.rds"
# )
```

## 5 Discussion

1.  [Why is the dataset of interest to you?](#q1)
2.  [What are the control and test conditions of the dataset?](#q2)
3.  [How many samples in each of the conditions of your dataset?](#q3)
4.  [Were there expression values that were not unique for specific genes? How did you handle these?](#q4)
5.  [Were there expression values that could not be mapped to current HUGO symbols?](#q5)
6.  [Were there any outliers in your dataset? How were they handled in the originating paper? How many outliers were removed?](#q6)
7.  [How did you handle replicates?](#q7)
8.  [What is the final coverage of your dataset?](#q8)

Additional "rough work"/trial code and reflections can be found [here](https://github.com/bcb420-2026/Christine_Cheng/wiki/Entry-5-%E2%80%90-Assignment-1).

## 6 References

```{r, include=FALSE, echo=FALSE}
# TODO: generate citation bib
print(citation("rmarkdown"), bibtex=TRUE)
print(citation("knitr"), bibtex=TRUE)
citation("BiocManager")
citation("GEOquery")
citation("dplyr")
print(citation("biomaRt"), bibtex=TRUE)

citation("SummarizedExperiment")  # TODO: if using data from getRNASeqData
citation("readr") # TODO: if reading raw counts file from getGEOSuppFiles
```
