---
title: "A1_Christine_Cheng"
author: "Christine En-Tse Cheng"
date: "02-10-2026"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: hide
bibliography: A1_references.bib    # TODO: insert the citations/bibliography file
link-citations: true            # this is for the citations/bibliography 
---
```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE, # suppressing warning messages
                      message = FALSE,
                      knitr.kable.max_rows = 6)

# install/attach packages
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

if (!require("GEOquery", quietly = TRUE))
  BiocManager::install("GEOquery")

library(SummarizedExperiment)


# read the data files - TODO: if needed

```

## Data Exploration
### Downloading the data set
```{r}
dataset_se <- GEOquery::getRNASeqData("GSE166605")
dataset_se
metadata(dataset_se)
rowData(dataset_se)
```
```{r}
# insert check for previously downloaded dataset in tempdir() or specified destdir
dataset_gse <- GEOquery::getGEO(GEO = "GSE166605", destdir = "~/projects/Christine_Cheng/A1_files", AnnotGPL = TRUE)
dataset_gse
```
```{r}
dataset_suppFiles <- GEOquery::getGEOSuppFiles("GSE166605", makeDirectory = FALSE, baseDir = "~/projects/Christine_Cheng/A1_files")
dataset_suppFiles
```

### Exploratory Statistics

### Gene ID mapping

### Data Trimming/Cleaning

## Data Normalization

## Differential Expression Analysis

## Discussion
1. Why is the dataset of interest to you?
2. What are the control and test conditions of the dataset?
3. How many samples in each of the conditions of your dataset?
4. Were there expression values that were not unique for specific genes? How did you handle these?
5. Were there expression values that could not be mapped to current HUGO symbols?
6. Were there any outliers in your dataset? How were they handled in the originating paper? How 7. many outliers were removed?
8. How did you handle replicates?
9. What is the final coverage of your dataset?

## References
---
nocite: '@*'
---

```{r}
# TODO: generate citation bib
citation("GEOquery")
print(citation("rmarkdown"), bibtex=TRUE)
```
