---
title: "A1 - Dataset selection, cleaning, normalization, and differential gene expression"
author: "Christine En-Tse Cheng"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
bibliography: A1_files/A1_references.bib    # TODO: insert the citations/bibliography file
link-citations: true            # this is for the citations/bibliography 
nocite: '@*'
---

```{r setup, include=FALSE}
# TODO: put back in yaml header above bibliography
# knit: (function(inputFile, encoding) {
#   rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs/A1") })

knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      knitr.kable.max_rows = 6)

# helper function to present formatted data table previews (default: 6 rows)
# TODO: remove
# kable_head <- function(x, n = 6, caption = NULL) {  
#   knitr::kable(utils::head(x, n), caption = caption)
# }

kable_head_helper <- function() {
  i <- 0
  function(x, n = 6, caption = NULL) {
    i <<- i + 1
    full_caption <- paste0("**Table ", i, ".** ", caption)
    knitr::kable(utils::head(x, n), caption = full_caption)
  }
}
kable_head <- kable_head_helper()
```

## 1 Introduction

TODO: introduce the dataset/paper, maybe briefly talk about how/why the dataset was chosen

This report describes the selection, data preparation, normalization, and differential expression analysis of a publicly available human gene expression dataset. TODO: The workflow follows best practices for RNA-seq analysis as outlined by (@dawadi_bench_2025)

### 1.1 Dataset Selection {#q1}

TODO: brief description The selected dataset is [GSE166605](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE166605) from the Gene Expression Omnibus (GEO). This RNA-seq dataset investigates TODO: [transcriptional responses of human CD14+ monocytes under experimentally defined conditions involving differential treatment exposure]. The dataset was chosen because TODO The selected dataset is GSE166605 from the Gene Expression Omnibus (GEO). This RNA-seq dataset investigates the transcriptional responses of human CD14+ monocytes. Specifically, it analyzes the effect of miR-155 overexpression compared to a negative control.

This dataset is of interest because miR-155 is a critical regulator of immune function and inflammation. Understanding its downstream targets in monocytes can reveal mechanisms of immune activation.

A detailed account of the dataset discovery process is documented separately [here](https://github.com/bcb420-2026/Christine_Cheng/wiki/Entry-4-%E2%80%90-Dataset-Selection-(2.3)).

### 1.2 Dataset Background

-   paper background(@olsson_mir-155-overexpressing_2022)

####  {#q2}

-   TODO: control and test conditions of dataset Organism: Homo sapiens

Tissue: CD14+ Monocytes

Control Condition: Negative Control (Neg_C)

Test Condition: miR-155 Overexpression (miR_155)

## 2 Data Exploration & Clean-up

The following sections describes initial steps in the analysis, from accessing the dataset to cleaning the data and performing preliminary exploratory analysis.

First, necessary libraries were installed (if necessary) and loaded into the R session.

```{r packages, include=FALSE}
# install/attach packages
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")


if (!require("GEOquery", quietly = TRUE))
  BiocManager::install("GEOquery")

if (!require("readr", quietly = TRUE))
  install.packages("readr")

if (!require("dplyr", quietly = TRUE))
  install.packages("dplyr")

if (!require("knitr", quietly = TRUE))
  install.packages("knitr")

if (!require("edgeR", quietly = TRUE))
  BiocManager::install("edgeR")

if (!require("biomaRt", quietly = TRUE))
  BiocManager::install("biomaRt")

if (!require("tibble", quietly = TRUE))
  install.packages("tibble")

if (!require("ggplot2", quietly = TRUE))
  install.packages("ggplot2")

if (!require("ComplexHeatmap", quietly = TRUE))
  BiocManager::install("ComplexHeatmap")

library(GEOquery)
library(readr)
library(dplyr)
library(knitr)
library(edgeR)
library(biomaRt)
library(tibble)
library(ggplot2)
library(ComplexHeatmap)
```

### 2.1 Obtaining the Data

The dataset was accessed and downloaded using `GEOquery` and the corresponding GEO series ID. The raw count data file was downloaded programmatically in R without manual file management, to ensure reproducibility of results. To ensure reproducibility while avoiding repeated downloads, files are only retrieved if not already present locally in designated data files directory.

TODO:make pretty: Below is a snapshot of the raw matrix:

```{r downloading}
GSE_ID <- "GSE166605" # constant for GEO series ID to reduce risk of accessing the wrong dataset
data_dir <- "A1_files"

# checking for existence of directory for storing data files
if (!dir.exists(data_dir)) 
  dir.create(data_dir)

counts_file <- file.path(data_dir, "GSE166605_D1-5_Neg_C_miR_155_RAW_counts.txt.gz")

# access raw files from GEO if the file(s) weren't previously downloaded
if (!file.exists(counts_file)) {
  GEOquery::getGEOSuppFiles(GSE_ID,
                            makeDirectory = FALSE,
                            baseDir = data_dir)
}

# read in raw count file: columns are samples and rows are genes
raw_counts <- readr::read_tsv(counts_file, show_col_types = FALSE)

# names(raw_counts)[1:min(12, ncol(raw_counts))]  # TODO: remove
# preview raw count data
kable_head(raw_counts, caption = "Preview of Raw Count Data from GEO") # TODO:make sure it renders decently/pretty
```

```{r}
# sanity check: gene IDs are Ensembl IDs
# all(grepl("^ENSG", raw_counts[[1]]))

# helper function to remove Ensembl IDs' version numbers:
strip_ensembl_version <- function(x) sub("\\..*$", "", x)

# Build a counts matrix (genes × samples)
counts <- raw_counts %>%
  dplyr::select(-`Gene name`) %>%
  dplyr::rename(ensembl_gene_id = GeneID) %>% # rename to be more descriptive and for easier ID mapping downstream
  as.data.frame()

counts$ensembl_gene_id <- strip_ensembl_version(counts$ensembl_gene_id)

ensembl <- biomaRt::useMart("ensembl", dataset = "hsapiens_gene_ensembl") # using latest Human Genes mart
ids_map <- biomaRt::getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = counts$ensembl_gene_id,
  mart = ensembl
)

mapped_counts <- counts %>%
  dplyr::left_join(ids_map, by = "ensembl_gene_id") %>%
  dplyr::select(ensembl_gene_id,hgnc_symbol, everything())
```

```{r}
# # found  duplicate Ensembl IDs
# duplicates <- mapped_counts %>%
#   dplyr::group_by(ensembl_gene_id) %>%
#   dplyr::filter(n() > 1) %>%
#   dplyr::arrange(ensembl_gene_id)
# 
# View(duplicates)
# 
# duplicate_summary <- mapped_counts %>%
#   dplyr::count(ensembl_gene_id) %>%
#   dplyr::filter(n > 1)
# 
# View(duplicate_summary)
# 
# id_conflicts <- mapped_counts %>%
#   dplyr::group_by(ensembl_gene_id, hgnc_symbol) %>%
#   dplyr::filter(n() > 1)
# 
# View(id_conflicts)
# 
# missing_gene <- setdiff(
#   unique(duplicate_summary$ensembl_gene_id), 
#   unique(id_conflicts$ensembl_gene_id)
# )
# 
# duplicates[duplicates$ensembl_gene_id == missing_gene, ] # same stripped Ensembl ID mapped to 2 different HGNC symbols!
# 
# raw_counts[raw_counts$GeneID == missing_gene, ] # raw counts doesn't have duplicated `missing_gene`!
# 
# # after running just:
# # mapped_counts <- counts %>%
# # +   dplyr::left_join(ids_map, by = "ensembl_gene_id")
# # mapped_counts already has the duplicated `missing_gene` -> BioMart mapping must be the issue
# ids_map[ids_map$ensembl_gene_id == missing_gene, ] # confirms that BioMart mapped two different HGNC symbols to the same Ensembl ID
```

```{r}
## FIX:
# 1. Collapse the map first so each Ensembl ID appears exactly once
ids_map_unique <- ids_map %>%
  dplyr::group_by(ensembl_gene_id) %>%
  dplyr::summarize(hgnc_symbol = paste(unique(hgnc_symbol), collapse = ", "))

# 2. Join with the collapsed map
mapped_counts_fixed <- counts %>%
  dplyr::left_join(ids_map_unique, by = "ensembl_gene_id")
```

```{r}
# duplicates1 <- mapped_counts_fixed %>%
#   dplyr::group_by(ensembl_gene_id) %>%
#   dplyr::filter(n() > 1) %>%
#   dplyr::arrange(ensembl_gene_id)
# 
# View(duplicates1)
# 
# duplicate_summary1 <- mapped_counts_fixed %>%
#   dplyr::count(ensembl_gene_id) %>%
#   dplyr::filter(n > 1)
# 
# View(duplicate_summary1)
# 
# id_conflicts1 <- mapped_counts_fixed %>%
#   dplyr::group_by(ensembl_gene_id, hgnc_symbol) %>%
#   dplyr::filter(n() > 1)
# 
# View(id_conflicts1)
# 
# nrow(duplicates1) == nrow(duplicate_summary1)*2
# nrow(duplicate_summary1)*2 == nrow(id_conflicts1)
# #fixed :)
```

```{r}
consistency_check <- mapped_counts_fixed %>%
  dplyr::group_by(ensembl_gene_id) %>%
  dplyr::summarize(
    row_count = n(),
    unique_symbols = n_distinct(hgnc_symbol),
    symbols_found = paste(unique(hgnc_symbol), collapse = "; ")
  ) %>%
  dplyr::filter(row_count > 1) %>% # Only look at the duplicates
  dplyr::filter(unique_symbols > 1) # Check if any duplicate Ensembl ID has more than 1 unique symbol

cat("Any more Ensembl IDs-HGNC symbols one-to-many mappings? ", nrow(consistency_check) > 0) 
```

```{r}
# Summing up expression values for duplicate Ensembl ID rows
final_counts <- mapped_counts_fixed %>%
  dplyr::group_by(ensembl_gene_id, hgnc_symbol) %>%
  dplyr::summarise(dplyr::across(tidyr::starts_with("D"), sum)) %>% # only sum sample columns (labels start /w 'D')
  dplyr::ungroup()

# # sanity check:
# length(unique(final_counts$ensembl_gene_id)) == nrow(final_counts)
```

```{r}
# BUT
hgnc_duplication <- length(unique(final_counts$hgnc_symbol)) == nrow(final_counts) # duplicate HGNC symbols still exist!
cat("All unique HGNC symbols? ", hgnc_duplication)
```

Since I plan on conducting differential expression analysis, I will not remove the rows with duplicated HGNC symbol as that may end up merging - Readthroughs: Two genes are transcribed together.

-   Paralogs/Homologs: Ensembl assigns distinct IDs to very similar regions (like ENSG...1 and ENSG...2), but the HGNC nomenclature considers them the same gene.

-   Pseudo-genes: A functional gene and its "broken" version might share a symbol but have separate Ensembl entries. [TODO: source]() which can result in not specific results. Instead the unique Ensembl IDs will be kept as a stable reference while the HGNC symbols (including duplicates) are kept as annotations.

```{r}
rownames(final_counts) <- final_counts$ensembl_gene_id # added rownames (succesfully)
# BUT, no matter which method, I can't get the rownames to stay:
# final_counts$ensembl_gene_id <- NULL
# final_counts <- dplyr::select(final_counts, !(ensembl_gene_id))
# final_counts <- final_counts %>% dplyr::select( -ensembl_gene_id)
# SOLUTION:
final_counts <- mapped_counts_fixed %>%
  dplyr::group_by(ensembl_gene_id, hgnc_symbol) %>%
  dplyr::summarise(dplyr::across(tidyr::starts_with("D"), sum)) %>% # results in tibble, that loses (temporary attached) rownames once another operation is applied on it
  dplyr::ungroup() %>%
  as.data.frame() # need to convert to standard data.frame here
```

```{r counts_objs}
# create objects for downstream DE analysis
count_mappings <- final_counts[, 1:2]

rownames(final_counts) <- final_counts$ensembl_gene_id
final_counts$ensembl_gene_id <- NULL
final_counts$hgnc_symbol <- NULL

# TODO: is this necessary???
# final_counts1 <- as.matrix(final_counts)
# storage.mode(final_counts1[, 2:ncol(final_counts1)]) <- "integer"
```

```{r metadata}
# Sample name format: D# CD14 control/treatment_name
samples <- colnames(final_counts)
cell_line <- unlist(lapply(samples,
                          FUN = function(x){unlist(strsplit(x,split = " "))[2]}))       

# TODO: remove
# treatment <- unlist(lapply(samples,FUN = 
#                              function(x){unlist(strsplit(x,split = " "))[3]}))     

is_control <- ifelse(grepl("Neg", samples, ignore.case = TRUE), "Control", "miR155")
condition = factor(is_control, levels = c("Control", "miR155"))

meta_df <- data.frame(samples, cell_line, condition)

kable_head(meta_df, n = 10, caption = "Sample Metadata")
```

```{r save_data}
# TODO: do i need?
# saveRDS(list(counts = final_counts,
#              meta = meta_df,
#              mappings = count_mappings),
#         file = "A1_files/counts_and_metadata.rds")
```

```{r}
# based on edgeR documentation, negative and NA count values are not permitted -> need to ensure it meets requirements
edger_precon <- all(sapply(final_counts, function(x) all(!is.na(x) & x >= 0 & x %% 1 == 0)))
cat("Final count matrix meets input preconditions? ", edger_precon)

dge <- edgeR::DGEList(final_counts)
dge$samples <- cbind(dge$samples, meta_df)
```

```{r filtering}
# Filter lowly expressed genes
# Purpose: Removing low-count genes improves statistical power by reducing the burden of multiple testing and removing "noisy" data that cannot reach significance anyway. 
design <- model.matrix(~ 0 + condition, data = meta_df)
# only creating the design contrast /w condition since all 10 samples were from the same cell line, thus control for batch effects and related is not necessary.

keep <- edgeR::filterByExpr(dge, design)
dge_filtered <- dge[keep, , keep.lib.sizes = FALSE]

filtered_percent <- round((1 - (nrow(dge_filtered) / nrow(dge))) * 100, digits = 2)
cat("Approximately ", filtered_percent, "% of genes were determined by the design matrix to be lowly expressed and thus removed for downstream analyses.")

# create a subset of the gene mappings
count_mappings_filtered <- count_mappings[count_mappings$ensembl_gene_id %in% rownames(dge_filtered$counts), ] 
```

```{r normalization}
dge_filtered <- edgeR::calcNormFactors(dge_filtered, method = "TMM")
dge_filtered$samples
```

```{r MDS_plot}
current_colors <- RColorBrewer::brewer.pal(n = length(unique(meta_df$condition)), "Set2")
names(current_colors) <- levels(factor(meta_df$condition))

plotMDS(dge_filtered,
  labels = meta_df$condition,
  col = current_colors[meta_df$condition],
  main = "MDS plot (colored by condition)"
)
legend("topright",
  legend = names(current_colors),
  col = current_colors,
  pch = 16,
  bty = "n"
)
# TODO: add caption
```

```{r log_CPM}
lcpm <- edgeR::cpm(dge_filtered, log = TRUE, prior.count = 1)
summary(as.vector(lcpm))
```

## 4 Differential Expression Analysis

I will now conduct differential expression analysis using the recommended quasi-likelihood pipeline from `edgeR` (@chen_reads_2016).

```{r dge_setup}
cat("Data design condition variable names: ", colnames(design))

# Contrasts
contr <- makeContrasts(
  `miR155_vs_Ctrl` = conditionmiR155 - conditionControl,
  levels = design
)

# Estimate dispersion and fit QL model
edge <- edgeR::estimateDisp(dge_filtered, design)
fit <- edgeR::glmQLFit(edge, design)
```

```{r bcv_plot}
plotBCV(edge)
# TODO: add caption
```
```{r dge_test}
miR155_vs_Ctrl <- edgeR::glmQLFTest(fit, contrast = contr)
dge_results <- edgeR::topTags(miR155_vs_Ctrl, n = Inf)$table %>% tibble::rownames_to_column("ensembl_id")
kable_head(dge_results, caption = "Preview of top DE genes") # TODO:make sure it renders decently/pretty
```

```{r volcano_plot}
volcano_df <- dge_results %>%
  dplyr::mutate(sig = FDR < 0.05)

ggplot2::ggplot(volcano_df,
                ggplot2::aes(x = logFC, y = -log10(PValue), color = sig)) +
  ggplot2::geom_point(alpha = 0.6) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "miR155 VS Neg. Ctrl",
    x = "log2 fold-change",
    y = "-log10(p-value)"
    # TODO: rename legend title
  )
# TODO: add caption
```
```{r volcano_plot2}
dge_results2 <- edgeR::topTags(miR155_vs_Ctrl, n = Inf)$table %>%
  tibble::rownames_to_column("ensembl_id") %>%
  dplyr::mutate(
    neglog10P = -log10(PValue),
    direction = case_when(
      FDR < 0.05 & logFC > 0 ~ "Up",
      FDR < 0.05 & logFC < 0 ~ "Down",
      TRUE ~ "Not sig"
    )
  )

ggplot2::ggplot(dge_results2, ggplot2::aes(x = logFC, y = neglog10P)) +
  ggplot2::geom_point(ggplot2::aes(color = direction), alpha = 0.7, size = 1.2) +
  ggplot2::scale_color_manual(values = c(
    "Up" = "red",
    "Down" = "blue",
    "Not sig" = "grey70"
  )) +
  ggplot2::geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "steelblue") +
  ggplot2::geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "steelblue") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "miR155 VS Neg. Ctrl",
    x = "log2 fold change (logFC)",
    y = "-log10(p-value)",
    color = NULL
  )
# TODO: add caption
```
```{r top_degs}
# Keep only the columns we need and ensure they are character
map_df <- count_mappings %>%
  dplyr::mutate(
    ensembl_id = as.character(ensembl_gene_id),
    hgnc_symbol = as.character(hgnc_symbol),
    .keep = "none"
  ) %>%
  dplyr::distinct(ensembl_id, .keep_all = TRUE)

# Join mapping onto the edgeR results table 
dge_results2_annotated <- dge_results2 %>%
  dplyr::mutate(ensembl_id = as.character(ensembl_id)) %>%
  dplyr::left_join(map_df, by = "ensembl_id") %>%
  dplyr::mutate(
    # Use HGNC when available, otherwise fall back to Ensembl
    display_name = ifelse(!is.na(hgnc_symbol) & hgnc_symbol != "",
                          hgnc_symbol,
                          ensembl_id)
  )

# Quick check: how many mapped?
mapped_percentage <- mean(!is.na(dge_results2_annotated$hgnc_symbol)) * 100
cat("Proportion (%) of maped DE results: ", mapped_percentage)

head(dge_results2_annotated[, c("ensembl_id", "hgnc_symbol", "display_name", "logFC", "FDR")])

# Top DEGs: all genes with FDR < 0.05
top_deg_hits <- dge_results2_annotated %>%
  dplyr::arrange(FDR, PValue) %>%
  dplyr::filter(FDR < 0.05)

# Ensembl IDs for subsetting the matrix
top_hits_ensembl <- top_deg_hits$ensembl_id

# Row labels we want to show on the heatmap
top_hits_labels <- top_deg_hits$display_name

head(top_deg_hits[, c("ensembl_id","display_name","logFC","FDR")])
```
```{r expression_matrix}
# logCPM from counts for visualization
# logCPM for visualization
logcpm <- edgeR::cpm(final_counts, log = TRUE, prior.count = 1)

# Keep only genes present in the expression matrix
keep <- top_hits_ensembl %in% rownames(logcpm)

hm_mat <- logcpm[top_hits_ensembl[keep], , drop = FALSE]

# Create labels aligned to hm_mat rows
hm_labels <- top_hits_labels[keep]

# IMPORTANT: avoid duplicate rownames (HGNC symbols can repeat)
# Make labels unique while preserving readability
rownames(hm_mat) <- make.unique(hm_labels)

# Z-score per gene across samples for pattern visibility
hm_z <- t(scale(t(hm_mat)))

dim(hm_z)
```
```{r heatmap_annotation}
meta_df$cell_line <- as.factor(meta_df$cell_line)
meta_df$condition <- as.factor(meta_df$condition)

# Set3 palettes (qualitative)
trtmnt_levels <- levels(meta_df$condition)
cell_levels  <- levels(meta_df$cell_line)

trtmnt_cols <- setNames(
  brewer.pal(max(3, length(trtmnt_levels)), "Set3")[seq_along(trtmnt_levels)],
  trtmnt_levels
)

cell_cols <- setNames(
  brewer.pal(max(3, length(cell_levels)), "Set2")[seq_along(cell_levels)],
  cell_levels
)

ha <- HeatmapAnnotation(
  treatment = meta_df$condition,
  cell_line  = meta_df$cell_line,
  col = list(
    treatment = trtmnt_cols,
    cell_line  = cell_cols
  ),
  annotation_legend_param = list(
    treatment = list(title = "Treatment"),
    cell_line  = list(title = "Cell line")
  )
)
```

```{r heatmap}
col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))

ComplexHeatmap::Heatmap(
  hm_z,
  col = col_fun,
  top_annotation = ha,
  show_row_names = FALSE,
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  row_title = paste0("Top ", nrow(hm_z), " genes (miR155 VS Neg. Ctrl)"),
  column_title = "Samples"
)

ComplexHeatmap::Heatmap(
  hm_z[1:25,], # TODO: Is this the correct way to add the top 25?
  col = col_fun,
  top_annotation = ha,
  show_row_names = TRUE,         # <-- now shows HGNC symbols
  row_names_gp = grid::gpar(fontsize = 8),
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  row_title = paste0("Top ", nrow(hm_z), " genes (miR155 VS Neg. Ctrl)"),
  column_title = "Samples"
)
# TODO: fix heatmap formatting + add captions
```

## 5 Discussion

1.  [Why is the dataset of interest to you?](#q1)
2.  [What are the control and test conditions of the dataset?](#q2)
3.  [How many samples in each of the conditions of your dataset?](#q3)
4.  [Were there expression values that were not unique for specific genes? How did you handle these?](#q4)
5.  [Were there expression values that could not be mapped to current HUGO symbols?](#q5)
6.  [Were there any outliers in your dataset? How were they handled in the originating paper? How many outliers were removed?](#q6)
7.  [How did you handle replicates?](#q7)
8.  [What is the final coverage of your dataset?](#q8)

Additional "rough work"/trial code and reflections can be found [here](https://github.com/bcb420-2026/Christine_Cheng/wiki/Entry-5-%E2%80%90-Assignment-1).

## 6 References

```{r, include=FALSE, echo=FALSE}
# TODO: generate citation bib
print(citation("rmarkdown"), bibtex=TRUE)
print(citation("knitr"), bibtex=TRUE)
citation("BiocManager")
citation("GEOquery")
citation("dplyr")
print(citation("biomaRt"), bibtex=TRUE)

# citation("SummarizedExperiment")  # TODO: if using data from getRNASeqData
citation("readr") # TODO: if reading raw counts file from getGEOSuppFiles
citation("edgeR")
citation("tibble")
citation("ggplot2")
print(citation("ComplexHeatmap"), bibtex = TRUE)
```
