---
title: "A1 - Dataset Selection, Cleaning, Normalization, and Differential Gene Expression"
author: "Christine En-Tse Cheng"
date: "2026-02-13"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs/A1") })
bibliography: A1_files/A1_references.bib
link-citations: true            # this is for the citations/bibliography 
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      knitr.kable.max_rows = 6)

# helper function to present formatted data table previews (default: 6 rows)
kable_head_helper <- function() {
  i <- 0
  function(x, n = 6, caption = NULL) {
    i <<- i + 1
    full_caption <- paste0("**Table ", i, ".** ", caption)
    knitr::kable(utils::head(x, n), caption = full_caption)
  }
}
kable_head <- kable_head_helper()
```

## 1 Introduction

This report describes the selection, data preparation, normalization, and differential expression analysis of a publicly available human gene expression dataset. The workflow follows best practices for RNA-seq analysis as outlined by (@dawadi_bench_2025).

### 1.1 Dataset Selection {#q1}

The selected dataset is [GSE166605](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE166605) from the Gene Expression Omnibus (GEO). This RNA-seq dataset investigates the transcriptional responses of healthy human CD14+ monocytes. Specifically, it analyzes the transcriptomic effect of microRNA *miR-155* overexpression compared to a negative control (@olsson_mir-155-overexpressing_2022).

#### {#q1}
This dataset is of interest because *miR-155* is a critical regulator of immune function and inflammation and has been shown to induce transcriptomic changes mimicking that of macrophages in rheumatoid arthritis (RA). Understanding its downstream targets in monocytes can reveal mechanisms of innate and adaptive immune responses in RA pathogenesis. Transcriptomic profiling of this dataset can reveal a link between microRNA dysregulation and pathogenic macrophage phenotypes. Findings regarding downstream *miR-155* targets has great potential for the identification of therapeutic checkpoints that drive chronic join inflammation (@olsson_mir-155-overexpressing_2022). I chose this dataset due to the applicability of this dataset's findings to potential therapeutics of a chronic disease, as this sort of analysis falls within my research interests of using computational tools for medical advances.

A detailed account of the dataset discovery process is documented separately [here](https://github.com/bcb420-2026/Christine_Cheng/wiki/Entry-4-%E2%80%90-Dataset-Selection-(2.3)).

### 1.2 Dataset Background
Here are sample high-level facts regarding the dataset and associated study:

- Organism: Homo sapiens

- Tissue: CD14+ Monocytes

- Control Condition: Negative Control (Neg_C)

- Test Condition: miR-155 over-expression (miR_155)

####  {#q2}

The study that generated the RNA-seq dataset studied the transcriptomic effects of *miR-155* overexpression in healthy monocytes against a negative control.The test condition group consists of healthy donor-derived preipheral blood CD14+ monocytes transfected with *miR-155* mimics (synthetic double-stranded RNA molecules that function as mature *miR-155*). The negative control samples consist of healthy donor-derived CD14+ monocytes transfected with a non-targeting control mimic (i.e., a scrambled/random sequence) (@olsson_mir-155-overexpressing_2022).

####  {#q3}

There are 5 samples for each of negative control and *miR-155* overexpression conditions in the dataset.

## 2 Data Exploration & Clean-up

The following sections describes initial steps in the analysis, from accessing the dataset to cleaning the data and performing preliminary exploratory analysis.

First, necessary libraries were installed (if necessary) and loaded into the R session.

```{r packages, include=FALSE}
# install/attach packages
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")


if (!require("GEOquery", quietly = TRUE)) {
  BiocManager::install("GEOquery")
  library(GEOquery)
}

if (!require("readr", quietly = TRUE)) {
  install.packages("readr")
  library(readr)
}

if (!require("dplyr", quietly = TRUE)) {
  install.packages("dplyr")
  library(dplyr)
}

if (!require("knitr", quietly = TRUE)) {
  install.packages("knitr")
  library(knitr)
}

if (!require("edgeR", quietly = TRUE)) {
  BiocManager::install("edgeR")
  library(edgeR) 
}

if (!require("biomaRt", quietly = TRUE)) {
  BiocManager::install("biomaRt")
  library(biomaRt) 
}

if (!require("tibble", quietly = TRUE)) {
  install.packages("tibble")
  library(tibble) 
}

if (!require("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
  library(ggplot2) 
}

if (!require("ComplexHeatmap", quietly = TRUE)) {
  BiocManager::install("ComplexHeatmap")
  library(ComplexHeatmap) 
}
```

### 2.1 Obtaining the Data

The dataset was accessed and downloaded using `GEOquery` and the corresponding GEO series ID. The raw count data file was downloaded programmatically in R without manual file management, to ensure reproducibility of results. To ensure reproducibility while avoiding repeated downloads, files are only retrieved if not already present locally in designated data files directory.

Below is a snapshot of the raw data from GEO:

```{r downloading}
GSE_ID <- "GSE166605" # constant for GEO series ID to reduce risk of accessing the wrong dataset
data_dir <- "A1_files"  # sub-directory for all A1 files

# checking for existence of directory for storing data files
if (!dir.exists(data_dir)) 
  dir.create(data_dir)

counts_file <- file.path(data_dir, "GSE166605_D1-5_Neg_C_miR_155_RAW_counts.txt.gz")

# access raw files from GEO if the file(s) weren't previously downloaded
if (!file.exists(counts_file)) {
  GEOquery::getGEOSuppFiles(GSE_ID,
                            makeDirectory = FALSE,
                            baseDir = data_dir)
}

# read in raw count file: columns are samples and rows are genes
raw_counts <- readr::read_tsv(counts_file, show_col_types = FALSE)

# preview raw count data
kable_head(raw_counts, caption = "Snapshot of Raw Count Data from GEO") 
```

### 2.2 Data Processing & Gene ID Mapping

Before using the raw counts for downstream analyses, I performed some basic processing and data cleaning. From the raw counts matrix, the 'Gene name' column was dropped and the Ensembl ID version numbers were stripped from th 'GeneID' column (which was also renamed to ensembl_gene_id for increased variable descriptiveness and easier downstream ID mapping).

```{r pre_processing}
# helper function to remove Ensembl IDs' version numbers using regex:
strip_ensembl_version <- function(x) sub("\\..*$", "", x)

# build counts matrix
counts <- raw_counts %>%
  dplyr::select(-`Gene name`) %>%
  dplyr::rename(ensembl_gene_id = GeneID) %>%
  as.data.frame()

# strip Ensemble ID version numbers
counts$ensembl_gene_id <- strip_ensembl_version(counts$ensembl_gene_id) 

kable_head(counts, caption = "Snapshot of Processed Count Matrix") 
```

#### Mapping Ensembl IDs to HGNC Symbols

Although the original count matrix contains gene names that resemble HGNC gene symbols, they were re-mapped to represent the most up-to-date gene annotations between the Ensembl and HUGO databases:

```{r id_map}
ensembl <- biomaRt::useMart("ensembl", dataset = "hsapiens_gene_ensembl") # using latest Human Genes mart

# create mapping between Ensembl IDs and HGNC gene symbols
ids_map <- biomaRt::getBM(
  # retrieving both Ensembl IDs and HGNC symbols as attribute to ensure we can map back to Ensembl gene ID in the original dataset if needed at any point in downstream analyses
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = counts$ensembl_gene_id,
  mart = ensembl
)

kable_head(ids_map, caption = "Preview of gene ID mapping between Ensembl IDs and HUGO gene IDs") 
```

####  {#q4}

Upon exploring the mapping between gene IDs, duplicate Ensembl IDs were discovered. This likely resulted from the version number stripping step. Through various troubleshooting steps (detailed [here](https://github.com/bcb420-2026/Christine_Cheng/wiki/Entry-5-%E2%80%90-Assignment-1#user-content-Troubleshooting_Ensembl_IDs_duplication)), I was able to verify that the duplication originated from BioMart mapping 2 different HGNC symbols to the same stripped Ensembl ID.

To address the issue, I first collapsed the gene ID map so that each Ensembl ID only appears once. For Ensembl IDs mapped to more than one HUGO symbol, I concatenated the HGNC symbols together on the row corresponding to the Ensembl ID. The expression data matrix is then joined with this collapsed ID map. This way, gene mapping data isn't lost and the final mapped counts matrix has unique row names (which is necessary for downstream analyses).

```{r id_mapping}
# 1. Collapse ID map
ids_map_unique <- ids_map %>%
  dplyr::group_by(ensembl_gene_id) %>%
  dplyr::summarize(hgnc_symbol = paste(unique(hgnc_symbol), collapse = ", "))

# 2. Join counts matrix with the collapsed ID map
mapped_counts_fixed <- counts %>%
  dplyr::left_join(ids_map_unique, by = "ensembl_gene_id")
```

I then checked the final mapped counts to ensure there's no more mis-mapping between Ensemble and HGNC:

```{r}
consistency_check <- mapped_counts_fixed %>%
  dplyr::group_by(ensembl_gene_id) %>%
  dplyr::summarize(
    row_count = n(),
    unique_symbols = n_distinct(hgnc_symbol),
    symbols_found = paste(unique(hgnc_symbol), collapse = "; ")
  ) %>%
  dplyr::filter(row_count > 1) %>% # Only look at the duplicates
  dplyr::filter(unique_symbols > 1) # Check if any duplicate Ensembl ID has more than 1 unique symbol

cat("Any more Ensembl IDs-HGNC symbols one-to-many mappings? ", nrow(consistency_check) > 0) 
```

However, there were still some non-unique rows (i.e., duplicate Ensembl IDs that has a matching pair of duplicate HGNC symbols). I summed up the expression values for these rows, to preserve the total transcriptional output for those genes; followed by a quick sanity check to ensure all duplicates were properly removed:

```{r}
# summing up expression values for duplicate Ensembl ID rows
final_counts <- mapped_counts_fixed %>%
  dplyr::group_by(ensembl_gene_id, hgnc_symbol) %>%
  dplyr::summarise(dplyr::across(tidyr::starts_with("D"), sum)) %>% # only sum sample columns (labels start /w 'D')
  dplyr::ungroup()

# sanity check
no_duplicates <- length(unique(final_counts$ensembl_gene_id)) == nrow(final_counts)
cat("All unique Ensembl IDs in count matrix? ", no_duplicates)
```

####  {#q7a}

However, duplicate HGNC symbols still exist!

```{r}
hgnc_duplication <- length(unique(final_counts$hgnc_symbol)) == nrow(final_counts)
cat("All unique HGNC symbols? ", hgnc_duplication)
```

Since I plan on conducting differential expression analysis, I will not remove the rows with duplicated HGNC symbol as that may end up accidentally cause non-specific results, like merging **paralogs/homologs** that have similar (but non-identical) Ensembl IDs due to close proximity in the genome, but are annotated as the same gene by HGNC nomenclature.

####  {#q5}

Similarly, there are some Ensembl IDs that couldn't be mapped to a HGNC symbol through the current HUGO-Ensembl ID mapping.

```{r}
# check if any Ensembl IDs failed to map to a HGNC symbol
unmapped <- sum(is.na(final_counts$hgnc_symbol)) > 0
cat("All rows mapped to HGNC symbol? ", !unmapped)
```

Instead of removing any genes for both these cases, the unique Ensembl IDs will be kept as a stable reference while the HGNC symbols (including duplicates and un-mapped) are kept as annotations:

```{r counts_objs}
final_counts <- mapped_counts_fixed %>%
  dplyr::group_by(ensembl_gene_id, hgnc_symbol) %>%
  dplyr::summarise(dplyr::across(tidyr::starts_with("D"), sum)) %>%
  dplyr::ungroup() %>%
  as.data.frame()

kable_head(final_counts, caption = "Preview of Annotated Count Matrix") 
```

### 2.3 Exploratory Statistics

For exploring the data, I first dropped any rows that has 0 expression or NA expression across all samples, if such rows exist. Then I calculated summary statistics for each sample:

```{r exploratory_stats}
# drop any rows that have 0 expression or NA expression across all samples
final_counts <- final_counts %>%
  dplyr::filter(dplyr::if_any(-c(1, 2), ~ !is.na(.x) & .x != 0))

# calculate summary statistics per sample
sample_summary <- data.frame(
  Total_Reads = colSums(final_counts[,-c(1,2)]),
  Mean_Expression = colMeans(final_counts[,-c(1,2)]),
  Median_Expression = apply(final_counts[,-c(1,2)], 2, median)
)

kable_head(sample_summary, caption = "Summary Statistics per Sample")

# check for outliers
boxplot(log2(final_counts[,-c(1,2)] + 1), 
        las = 2, col = "skyblue", 
        main = "Log2 Counts Distribution",
        ylab = "Log2(Counts + 1)")
```

#### {#q6}
Based on the summary table and boxplot, it appears there's a relatively consistent across all 10 samples, which matches with the fact that the original study didn't remove any outliers.

####  {#q7b}

```{r dupl_samples}
# check for expression value duplicates across samples
dup_samples <- final_counts[, c(FALSE, FALSE, duplicated(as.list(final_counts[, 3:ncol(final_counts)])))]
cat("\nUnique expression values between samples? ", ncol(dup_samples) == 0)
```

## 3 Data filtering and normalization

This next section describes the steps taken to prepare the dataset for differential expression analysis, specifically using `edgeR`.

### 3.1 Creating inputs for `edgeR`

Based on `edgeR` documentation, negative and `NA` count values are not permitted (these types of values also don't make sense biologically), so the count matrix was checked to ensure `edgeR` input preconditions are met:

```{r edgeR_precond}
# create objects for downstream DE analysis (meeting edgeR input preconditions)

count_mappings <- final_counts[, 1:2] # clean map between Ensembl IDs and HGNC symbols

# setting up count matrix to match expected row names and column names
rownames(final_counts) <- final_counts$ensembl_gene_id
final_counts$ensembl_gene_id <- NULL
final_counts$hgnc_symbol <- NULL

valid_counts <- all(sapply(final_counts, is.numeric))
cat("Valid count matrix values? ", valid_counts)

# check that no expression values are negative or NA
edger_precon <- all(sapply(final_counts, function(x) all(!is.na(x) & x >= 0 & x %% 1 == 0)))
cat("\nFinal count matrix meets input preconditions? ", edger_precon)
```

### 3.2 Preparing Metadata

In addition to the cleaned counts matrix and gene ID mapping, metadata outlining the dataset samples, cell line(s), and condition(s) is also needed:

```{r metadata}
# Sample name format: D# CD14 control/treatment_name
samples <- colnames(final_counts)
cell_line <- unlist(lapply(samples,
                          FUN = function(x){unlist(strsplit(x,split = " "))[2]}))       

is_control <- ifelse(grepl("Neg", samples, ignore.case = TRUE), "Control", "miR155")
condition = factor(is_control, levels = c("Control", "miR155"))

meta_df <- data.frame(samples, cell_line, condition)

kable_head(meta_df, n = 10, caption = "Sample Metadata")
```

### 3.3 Data filtering

Using an `edgeR` data container and package-level functions, lowly expressed genes were filtered out based on threholds set by `edgeR` based on the design matrix of the dataset (@chen_reads_2016). The purpose of this step is to improve statistical power by reducing the burden of multiple testing and removing "noisy" data that cannot reach significance anyway.

```{r filtering}
# create edgeR data container
dge <- edgeR::DGEList(final_counts)
dge$samples <- cbind(dge$samples, meta_df)  # adding metadata to sample data

# create dataset design matrix
  # only creating the design contrast /w condition since all 10 samples were from the same cell line, thus control for batch effects and related is not necessary
design <- model.matrix(~ 0 + condition, data = meta_df)

# filter out lowly expressed genes and recalculate library sizes based on remaining genes
keep <- edgeR::filterByExpr(dge, design)
dge_filtered <- dge[keep, , keep.lib.sizes = FALSE]

filtered_percent <- round((1 - (nrow(dge_filtered) / nrow(dge))) * 100, digits = 2)
cat("Approximately ", filtered_percent, "% of genes had low expression and thus removed for downstream analyses.")

kable_head(dge_filtered$counts, caption = "Snapshot of Filtered Count Matrix") 

# create a subset of the gene mappings
count_mappings_filtered <- count_mappings[count_mappings$ensembl_gene_id %in% rownames(dge_filtered$counts), ] 
```

### 3.4 Apply Normalization

Prior to differential expression analysis, I performed Trimmed Mean of M-values (TMM) normalization on the cleaned and filtered count data (@robinson_scaling_2010). I used the TMM method via `edgeR` since it is preferred for RNA-seq as it accounts for compositional differences between libraries (@dawadi_bench_2025). Below are two plots, first before normalization and second after.

```{r normalization}
# plot of data pre-normalization
log_counts_pre <- edgeR::cpm(dge, log = TRUE)
plot(density(log_counts_pre[,1]), col = "red", main = "Density: Pre vs Post Normalization", lwd = 2)
for(i in 2:ncol(log_counts_pre)){ lines(density(log_counts_pre[,i]), col = "red", lty = 2) }
log_counts_pre

dge_filtered <- edgeR::calcNormFactors(dge_filtered, method = "TMM")
dge_filtered$samples

# plot of data post-normalization
log_counts_post <- edgeR::cpm(dge_filtered, log = TRUE)
for(i in 1:ncol(log_counts_post)){ lines(density(log_counts_post[,i]), col = "blue") }
legend("topright", legend = c("Pre-Norm", "Post-Norm"), col = c("red", "blue"), lty = 1)
log_counts_post
```

```{r MDS_plot}
current_colors <- RColorBrewer::brewer.pal(n = length(unique(meta_df$condition)), "Set2")
names(current_colors) <- levels(factor(meta_df$condition))

plotMDS(dge_filtered,
  labels = meta_df$condition,
  col = current_colors[meta_df$condition],
  main = "MDS plot (colored by condition)"
)
legend("topright",
  legend = names(current_colors),
  col = current_colors,
  pch = 16,
  bty = "n"
)
```

```{r log_CPM}
lcpm <- edgeR::cpm(dge_filtered, log = TRUE, prior.count = 1)
summary(as.vector(lcpm))
```

## 4 Differential Expression Analysis

Then I conducted differential expression analysis using the recommended quasi-likelihood pipeline from `edgeR` (@chen_reads_2016).

```{r dge_setup}
cat("Data design condition variable names: ", colnames(design))

# Contrasts
contr <- makeContrasts(
  `miR155_vs_Ctrl` = conditionmiR155 - conditionControl,
  levels = design
)

# Estimate dispersion and fit QL model
edge <- edgeR::estimateDisp(dge_filtered, design)
fit <- edgeR::glmQLFit(edge, design)
```

```{r bcv_plot}
plotBCV(edge)
```

```{r dge_test}
miR155_vs_Ctrl <- edgeR::glmQLFTest(fit, contrast = contr)
dge_results <- edgeR::topTags(miR155_vs_Ctrl, n = Inf)$table %>% tibble::rownames_to_column("ensembl_id")
kable_head(dge_results, caption = "Preview of top DE genes")
```

```{r volcano_plot}
volcano_df <- dge_results %>%
  dplyr::mutate(sig = FDR < 0.05)

ggplot2::ggplot(volcano_df,
                ggplot2::aes(x = logFC, y = -log10(PValue), color = sig)) +
  ggplot2::geom_point(alpha = 0.6) +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "miR155 VS Neg. Ctrl",
    x = "log2 fold-change",
    y = "-log10(p-value)"
  )
```

```{r volcano_plot2}
dge_results2 <- edgeR::topTags(miR155_vs_Ctrl, n = Inf)$table %>%
  tibble::rownames_to_column("ensembl_id") %>%
  dplyr::mutate(
    neglog10P = -log10(PValue),
    direction = case_when(
      FDR < 0.05 & logFC > 0 ~ "Up",
      FDR < 0.05 & logFC < 0 ~ "Down",
      TRUE ~ "Not sig"
    )
  )

ggplot2::ggplot(dge_results2, ggplot2::aes(x = logFC, y = neglog10P)) +
  ggplot2::geom_point(ggplot2::aes(color = direction), alpha = 0.7, size = 1.2) +
  ggplot2::scale_color_manual(values = c(
    "Up" = "red",
    "Down" = "blue",
    "Not sig" = "grey70"
  )) +
  ggplot2::geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "steelblue") +
  ggplot2::geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "steelblue") +
  ggplot2::theme_minimal() +
  ggplot2::labs(
    title = "miR155 VS Neg. Ctrl",
    x = "log2 fold change (logFC)",
    y = "-log10(p-value)",
    color = NULL
  )
```

```{r top_degs}
# Keep only the columns we need and ensure they are character
map_df <- count_mappings %>%
  dplyr::mutate(
    ensembl_id = as.character(ensembl_gene_id),
    hgnc_symbol = as.character(hgnc_symbol),
    .keep = "none"
  ) %>%
  dplyr::distinct(ensembl_id, .keep_all = TRUE)

# Join mapping onto the edgeR results table 
dge_results2_annotated <- dge_results2 %>%
  dplyr::mutate(ensembl_id = as.character(ensembl_id)) %>%
  dplyr::left_join(map_df, by = "ensembl_id") %>%
  dplyr::mutate(
    # Use HGNC when available, otherwise fall back to Ensembl
    display_name = ifelse(!is.na(hgnc_symbol) & hgnc_symbol != "",
                          hgnc_symbol,
                          ensembl_id)
  )

# Quick check: how many mapped?
mapped_percentage <- mean(!is.na(dge_results2_annotated$hgnc_symbol)) * 100
cat("Proportion (%) of maped DE results: ", mapped_percentage)

head(dge_results2_annotated[, c("ensembl_id", "hgnc_symbol", "display_name", "logFC", "FDR")])

# Top DEGs: all genes with FDR < 0.05
top_deg_hits <- dge_results2_annotated %>%
  dplyr::arrange(FDR, PValue) %>%
  dplyr::filter(FDR < 0.05)

# Ensembl IDs for subsetting the matrix
top_hits_ensembl <- top_deg_hits$ensembl_id

# Row labels we want to show on the heatmap
top_hits_labels <- top_deg_hits$display_name

head(top_deg_hits[, c("ensembl_id","display_name","logFC","FDR")])
```

```{r expression_matrix}
# logCPM from counts for visualization
# logCPM for visualization
logcpm <- edgeR::cpm(final_counts, log = TRUE, prior.count = 1)

# Keep only genes present in the expression matrix
keep <- top_hits_ensembl %in% rownames(logcpm)

hm_mat <- logcpm[top_hits_ensembl[keep], , drop = FALSE]

# Create labels aligned to hm_mat rows
hm_labels <- top_hits_labels[keep]

# IMPORTANT: avoid duplicate rownames (HGNC symbols can repeat)
# Make labels unique while preserving readability
rownames(hm_mat) <- make.unique(hm_labels)

# Z-score per gene across samples for pattern visibility
hm_z <- t(scale(t(hm_mat)))

dim(hm_z)
```

```{r heatmap_annotation}
meta_df$cell_line <- as.factor(meta_df$cell_line)
meta_df$condition <- as.factor(meta_df$condition)

# Set3 palettes (qualitative)
trtmnt_levels <- levels(meta_df$condition)
cell_levels  <- levels(meta_df$cell_line)

trtmnt_cols <- setNames(
  RColorBrewer::brewer.pal(max(3, length(trtmnt_levels)), "Set3")[seq_along(trtmnt_levels)],
  trtmnt_levels
)

cell_cols <- setNames(
  RColorBrewer::brewer.pal(max(3, length(cell_levels)), "Set2")[seq_along(cell_levels)],
  cell_levels
)

ha <- HeatmapAnnotation(
  treatment = meta_df$condition,
  cell_line  = meta_df$cell_line,
  col = list(
    treatment = trtmnt_cols,
    cell_line  = cell_cols
  ),
  annotation_legend_param = list(
    treatment = list(title = "Treatment"),
    cell_line  = list(title = "Cell line")
  )
)
```

```{r heatmap}
col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))

ComplexHeatmap::Heatmap(
  hm_z,
  col = col_fun,
  top_annotation = ha,
  show_row_names = FALSE,
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  row_title = paste0("Top ", nrow(hm_z), " genes (miR155 VS Neg. Ctrl)"),
  column_title = "Samples"
)

ComplexHeatmap::Heatmap(
  hm_z[1:25,],
  col = col_fun,
  top_annotation = ha,
  show_row_names = TRUE,         # <-- now shows HGNC symbols
  row_names_gp = grid::gpar(fontsize = 8),
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  row_title = paste0("Top ", nrow(hm_z), " genes (miR155 VS Neg. Ctrl)"),
  column_title = "Samples"
)
```

#### {#q8}
Final coverage of the dataset:

```{r}
cat("The final dataset, after filtering lowly expressed genes, contains ", nrow(dge_filtered) , " genes across 10 samples")
```
This can be considered a robust coverage of the human protein-coding transcriptome suitable for pathway analysis.

## 5 Discussion

1.  [Why is the dataset of interest to you?](#q1)
2.  [What are the control and test conditions of the dataset?](#q2)
3.  [How many samples in each of the conditions of your dataset?](#q3)
4.  [Were there expression values that were not unique for specific genes? How did you handle these?](#q4)
5.  [Were there expression values that could not be mapped to current HUGO symbols?](#q5)
6.  [Were there any outliers in your dataset? How were they handled in the originating paper? How many outliers were removed?](#q6)

In the original paper, there is no explicit mention of outliers being removed from the dataset.

7.  How did you handle replicates? handled in [gene ID duplications](#q7a) and [expression duplicates across samples](#q7b)
8.  [What is the final coverage of your dataset?](#q8)

Additional "rough work"/trial code and reflections can be found [here](https://github.com/bcb420-2026/Christine_Cheng/wiki/Entry-5-%E2%80%90-Assignment-1).

## 6 References
