---
title: "A1 - Dataset selection, cleaning, normalization, and differential gene expression"
author: "Christine En-Tse Cheng"
date: "`r Sys.Date()"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs/A1") })
bibliography: A1_files/A1_references.bib    # TODO: insert the citations/bibliography file
link-citations: true            # this is for the citations/bibliography 
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      knitr.kable.max_rows = 6)

# helper function to present formatted data table previews (default: 6 rows)
# TODO: remove
# kable_head <- function(x, n = 6, caption = NULL) {  
#   knitr::kable(utils::head(x, n), caption = caption)
# }

kable_head_helper <- function() {
  i <- 0
  function(x, n = 6, caption = NULL) {
    i <<- i + 1
    full_caption <- paste0("**Table ", i, ".** ", caption)
    knitr::kable(utils::head(x, n), caption = full_caption)
  }
}
kable_head <- kable_head_helper()
```

## 1 Introduction

TODO: introduce the dataset/paper, maybe briefly talk about how/why the dataset was chosen

This report describes the selection, data preparation, normalization, and differential expression analysis of a publicly available human gene expression dataset. TODO: The workflow follows best practices for RNA-seq analysis as outlined by Dawadi et al. (2025) [@dawadi_bench_2025]

### 1.1 Dataset Selection {#q1}

TODO: brief description
The selected dataset is [GSE166605](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE166605) from the Gene Expression Omnibus (GEO). This RNA-seq dataset investigates TODO: [transcriptional responses of human CD14+ monocytes under experimentally defined conditions involving differential treatment exposure]. The dataset was chosen because TODO
The selected dataset is GSE166605 from the Gene Expression Omnibus (GEO). This RNA-seq dataset investigates the transcriptional responses of human CD14+ monocytes. Specifically, it analyzes the effect of miR-155 overexpression compared to a negative control.

This dataset is of interest because miR-155 is a critical regulator of immune function and inflammation. Understanding its downstream targets in monocytes can reveal mechanisms of immune activation.

A detailed account of the dataset discovery process is documented separately [here](https://github.com/bcb420-2026/Christine_Cheng/wiki/Entry-4-%E2%80%90-Dataset-Selection-(2.3)).

### 1.2 Dataset Background

-   paper background(@olsson_mir-155-overexpressing_2022)

####  {#q2}

-   TODO: control and test conditions of dataset
Organism: Homo sapiens

Tissue: CD14+ Monocytes

Control Condition: Negative Control (Neg_C)

Test Condition: miR-155 Overexpression (miR_155)

## 2 Data Exploration & Clean-up

The following sections describes initial steps in the analysis, from accessing the dataset to cleaning the data and performing preliminary exploratory analysis.

First, necessary libraries were installed (if necessary) and loaded into the R session.

```{r packages, include=FALSE}
# install/attach packages
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")


if (!require("GEOquery", quietly = TRUE))
  BiocManager::install("GEOquery")

if (!require("readr", quietly = TRUE))
  install.packages("readr")

if (!require("dplyr", quietly = TRUE))
  install.packages("dplyr")

if (!require("knitr", quietly = TRUE))
  install.packages("knitr")

if (!require("biomaRt", quietly = TRUE))
  BiocManager::install("biomaRt")

library(GEOquery)
library(readr)
library(dplyr)
library(knitr)
library(biomaRt)
library(readr)
```

### 2.1 Obtaining the Data

The dataset was accessed and downloaded using `GEOquery` and the corresponding GEO series ID. The raw count data file was downloaded programmatically in R without manual file management, to ensure reproducibility of results. To ensure reproducibility while avoiding repeated downloads, files are only retrieved if not already present locally in designated data files directory.

TODO:make pretty: Below is a snapshot of the raw matrix:

```{r downloading}
GSE_ID <- "GSE166605" # constant for GEO series ID to reduce risk of accessing the wrong dataset
data_dir <- "A1_files"

# checking for existence of directory for storing data files
if (!dir.exists(data_dir)) 
  dir.create(data_dir)

counts_file <- file.path(data_dir, "GSE166605_D1-5_Neg_C_miR_155_RAW_counts.txt.gz")

# access raw files from GEO if the file(s) weren't previously downloaded
if (!file.exists(counts_file)) {
  GEOquery::getGEOSuppFiles(GSE_ID,
                            makeDirectory = FALSE,
                            baseDir = data_dir)
}

# read in raw count file: columns are samples and rows are genes
raw_counts <- readr::read_tsv(counts_file, show_col_types = FALSE)

# names(raw_counts)[1:min(12, ncol(raw_counts))]  # TODO: remove
# preview raw count data
kable_head(raw_counts, caption = "Preview of Raw Count Data from GEO") # TODO:make sure it renders decently/pretty
```

The raw counts file was additionally processed to remove gene IDs and gene names (as they will be manually mapped later on in the workflow TODO: check if paper describes how they annotated the gene names > then decide if I should keep it or not), such that only raw expression values remained: TODO: display `counts`

TODO: Add comments explaining:
- Why raw counts are required for edgeR
- Why integers are necessary (NB distribution assumption)(?)

```{r pre_processing}
# processing count file to only contain expression values (remove GeneID and Gene Name columns)
counts <- raw_counts %>%
  dplyr::select(starts_with("D")) %>% # based on GEO accession page, all sample names start with 'D'
  as.data.frame()

rownames(counts) <- raw_counts$GeneID

counts <- as.matrix(counts) # converted to matrix for compatibility with DE analysis packages

# check that all values in the counts matrix are *mathematically* integers (even if the storage type isn't) before converting storage type of count values to integers
stopifnot(all(counts %% 1 == 0))
storage.mode(counts) <- "integer"

counts # TODO: display `counts`
```

TODO:remove??? Parsing count data to make metadata for design matrix.

```{r metadata}
# # We parse the column names to determine groups.
# # Expected format based on filename: D1_Neg_C_... or similar.
# sample_names <- colnames(counts_matrix)
# 
# # Regex logic: Look for "Neg" for control, "miR" for treatment
# conditions <- ifelse(grepl("Neg", sample_names, ignore.case = TRUE), "Control", "miR155")
# 
# meta_df <- data.frame(
#   Sample = sample_names,
#   Condition = factor(conditions, levels = c("Control", "miR155"))
# )

# Sample name format: D# CD14 control/treatment_name
samples <- colnames(counts)[1:ncol(counts)]
cell_line  <- unlist(lapply(samples,FUN =
                             function(x){unlist(strsplit(x,split = " "))[2]}))       
treatment <- unlist(lapply(samples,FUN = 
                             function(x){unlist(strsplit(x,split = " "))[3]}))     
meta_df <- data.frame(samples, cell_line, treatment)

kable_head(meta_df, caption = "Sample Metadata Preview")
```

### 2.2 Gene ID mapping

```{r id_mapping}
# helper function to remove Ensembl IDs' version numbers:
strip_ensembl_version <- function(x) sub("\\..*$", "", x)

counts <- data.frame(ensembl_gene_id = rownames(counts),
                            counts)

counts$ensembl_gene_id <- strip_ensembl_version(counts$ensembl_gene_id)
# TODO: check for duplicate ensembl IDs now that version numbers have been stripped

ensembl <- biomaRt::useMart("ensembl", dataset = "hsapiens_gene_ensembl") # using latest Human Genes mart
map <- biomaRt::getBM(
  #attributes should include both ensembl gene ID and hgnc symbol to ensure we can map back to the ensembl gene in the original data from hgnc at any point in our analysis 
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = counts$ensembl_gene_id,
  mart = ensembl
)

# TODO: remove?
# kable_head(map, 10, paste(": mapping preview"))

counts_mapped <- counts %>%
  dplyr::left_join(map, by = "ensembl_gene_id") %>%
  dplyr::select(ensembl_gene_id,hgnc_symbol, everything())

kable_head(counts_mapped[, 1:min(8, ncol(counts_mapped))], caption = "Mapped Counts Preview")

# Drop rows with empty/NA HGNC symbols
# TODO: fill in number of dropped rows with missing HGNC symbols {#q5}
counts_mapped[!(is.na(counts_mapped$hgnc_symbol) | counts_mapped$hgnc_symbol == ""), ]

counts_mapped <- counts_mapped[!(is.na(counts_mapped$hgnc_symbol) | counts_mapped$hgnc_symbol == ""), ]

  # Aggregate duplicates: for HGNC symbols appearing more than once, sum counts across duplicates (common, simple approach for gene‑level DGE). State the alternative (choose the row with highest average count) and why summation is acceptable for gene‑level DGE. Then set rownames to unique HGNC. 
  # Resolution: If multiple rows map to the same HUGO symbol (common with splice variants or probes), we sum the counts. This preserves the total transcriptional output for that gene.
#   # AGGREGATION: Sum counts for duplicate HGNC symbols
# # This handles the "Many-to-One" mapping problem
# final_counts <- mapped_data %>%
#   dplyr::select(-id) %>% # Remove original ID
#   dplyr::group_by(hgnc_symbol) %>%
#   dplyr::summarise(across(everything(), sum)) %>%
#   as.data.frame()

  # In your text, explain how you handled: unmapped IDs, multi‑mapping, duplicate symbols — these directly answer Q4–Q5. For biomaRt/Ensembl usage notes, reference the vignette.

# # Set rownames
# rownames(final_counts) <- final_counts$hgnc_symbol
# final_counts$hgnc_symbol <- NULL
# final_matrix <- as.matrix(final_counts)
# 
# cat("Original genes:", nrow(counts_matrix), "\n")
# cat("Mapped & Unique HUGO genes:", nrow(final_matrix), "\n")
```

### 2.3 Exploratory Statistics {#q3}

At first glance, the dataset contains TODO

-   evaluate sequencing depth and library size comparisons across samples??

```{r}
dim(counts)
summary(colSums(counts))

# RNA-seq counts are depth dependent.
# Library sizes and zero‑count genes:
lib_sizes <- colSums(counts)
summary(lib_sizes)
# TODO: Add:
  # Boxplot of library sizes: boxplot of log2(lib_sizes)
  # histogram of per‑gene zero counts
  # Barplot with sample labels

# TODO: Quick MDS (after we build a DGEList in next phase) colored by treatment and by donor (paired design sanity check). We’ll come back to this once normalization factors are in (Phase 5).
```
TODO: Interpret:
- Are depths comparable?
- Any extreme sample?

```{r}
# # initial logCPM
# dge <- edgeR::DGEList(counts)
# logcpm <- edgeR::cpm(dge, log = TRUE, prior.count = 1)
# # TODO: Plot: Density plot per sample
```
TODO: Explain:
- Log transformation stabilizes variance for visualization (Dawadi et al., 2025).

### 2.4 Data Trimming/Cleaning

```{r}
# design <- model.matrix(~ treatment, data = meta)
# 
# keep <- edgeR::filterByExpr(dge, design)
# dge <- dge[keep, , keep.lib.sizes = FALSE]
```
TODO:
Explain clearly:
- Why low-count genes are removed
- How filtering improves dispersion estimation
- Cite Dawadi paper discussion on low-count instability

Report:
- Number of genes before filtering
- After filtering

####  {#q6}

TODO: filtering low counts
I filtered genes with low expression to improve statistical power and reduce noise.

```{r}
# dge <- edgeR::DGEList(counts = final_matrix, group = metadata$Condition)
# 
# # Filter: Keep genes with > 1 CPM in at least 2 samples (smallest group size)
# keep <- edgeR::filterByExpr(dge, group = metadata$Condition)
# dge <- dge[keep, , keep.lib.sizes = FALSE]
# 
# cat("Genes remaining after filtering:", nrow(dge$counts))
```

####  {#q7}

```{r}
# # QC, filtering and identifier mapping
# # This chapter performs shared QC steps used by both limma-voom and edgeR and DEseq2.
# 
# obj <- readRDS("data/d1_counts_and_meta.rds")
# counts <- obj$counts
# meta <- obj$meta
# mappings <- obj$mappings
# 
# # Create a DGEList (edgeR data container)
# dge <- edgeR::DGEList(counts = counts)
# 
# # Attach metadata
# # (not required, but helpful)
# dge$samples <- cbind(dge$samples, meta)
# 
# dge$samples
# 
# # filter lowly expressed genes
# design <- model.matrix(~ 0 +  treatment+ cellline, data = meta)
# 
# keep <- edgeR::filterByExpr(dge, design)
# dge_f <- dge[keep, , keep.lib.sizes = FALSE]
# 
# dim(dge)
```

## 3 Data Normalization
Prior to differential expression analysis, I performed Trimmed Mean of M-values (TMM) normalization (@robinson_scaling_2010) on the cleaned count data. 3.1`

### 3.1 Normalization with TMM

TODO: I used the TMM (Trimmed Mean of M-values) method via `edgeR`. TMM is preferred for RNA-seq as it accounts for compositional differences between libraries (Dawadi et al., 2025).

```{r}
# # TMM normalization
# dge_f <- edgeR::calcNormFactors(dge_f, method = "TMM")
# # dge_f$samples
# # TODO: Include table of normalization factors.
```
TODO: explain: Why TMM?
- Adjusts library composition bias
- Trims extreme genes
- Good when majority of genes are not DE
- Recommended for between-sample comparison
- Consistent with Dawadi et al.

```{r}
# # exploratory MDS plot > to identify outliers (?)
# current_colors <- RColorBrewer::brewer.pal(n = length(unique(meta$treatment)), "Set2")
# names(current_colors) <- levels(factor(meta$treatment))
# 
# plotMDS(dge_f,
#   labels = meta$treatment,
#   col = current_colors[meta$treatment],
#   main = "MDS plot (colored by treatment)"
# ) # TODO: (if not already) color by condition

# legend("topright",
#   legend = names(current_colors),
#   col = current_colors,
#   pch = 16,
#   bty = "n"
# )
# 
# # " " " - coloured by cell line
# current_colors_cellline <- RColorBrewer::brewer.pal(n = length(unique(meta$cellline)), "Set1")
# names(current_colors_cellline) <- levels(factor(meta$cellline))
# 
# plotMDS(dge_f,
#   labels = meta$cellline,
#   col = current_colors_cellline[meta$cellline],
#   main = "MDS plot (colored by cellline)"
# )
# legend("topright",
#   legend = names(current_colors_cellline),
#   col = current_colors_cellline,
#   pch = 16,
#   bty = "n"
# )
```
TODO:
Interpret:

Do replicates cluster?

Does treatment separate?

Any outliers?

If outlier:

Check paper

Justify removal

Remove explicitly and document

This answers:

Outlier question

Replicate handling

Model design justification

```{r}
# # Calculate CPM (Counts Per Million) for visualizatio
# lcpm <- edgeR::cpm(dge_f, log = TRUE, prior.count = 1)
# summary(as.vector(lcpm))
```

```{r}
# # Visualization of normalization
# par(mfrow = c(1, 2))
# # Pre-normalization (using raw log CPM for comparison)
# raw_cpm <- edgeR::cpm(final_matrix, log = TRUE)
# boxplot(raw_cpm, las = 2, main = "Pre-Normalization (Log CPM)", 
#         col = as.numeric(metadata$Condition))
# 
# # Post-normalization
# boxplot(cpm_log, las = 2, main = "Post-TMM Normalization", 
#         col = as.numeric(metadata$Condition))
# legend("topright", legend = levels(metadata$Condition), fill = 1:2, cex = 0.7)
```
TODO: Interpretation: The boxplots verify that TMM normalization has centered the expression distributions, making samples comparable for differential analysis.

```{r}
# saveRDS(list(dge_f = dge_f, 
#              meta = meta, 
#              design = design, 
#              lcpm = lcpm, 
#              mappings = mappings),
#         file = "data/d1_qc_objects.rds")
```

### 3.2 Sample Validation (MDS/PCA)



```{r}
# library(dplyr)
# library(stringr)
# 
# # ---- 1) Set these to your actual column names ----
# # Example: raw_counts has columns "ensembl_id" and "gene_symbol"
# #          counts_mapped has columns "gene_id" and "gene_name"
# id_col_raw    <- "GeneID"
# name_col_raw  <- "Gene name"
# id_col_map    <- "ensembl_gene_id"
# name_col_map  <- "hgnc_symbol"
# 
# # ---- 2) Keep only the two columns and standardize names ----
# raw_pairs <- raw_counts %>%
#   transmute(
#     gene_id   = .data[[id_col_raw]],
#     gene_name = .data[[name_col_raw]]
#   )
# 
# map_pairs <- counts_mapped %>%
#   transmute(
#     gene_id   = .data[[id_col_map]],
#     gene_name = .data[[name_col_map]]
#   )
# 
# # ---- (Optional but helpful) 3) Normalize text to avoid trivial mismatches ----
# normalize <- function(x) {
#   x %>%
#     as.character() %>%
#     str_trim() %>%
#     str_squish()
# }
# raw_pairs <- raw_pairs %>%
#   mutate(across(everything(), normalize))
# map_pairs <- map_pairs %>%
#   mutate(across(everything(), normalize))
# 
# # If case-insensitive matching is desired for names (IDs usually case-sensitive):
# # raw_pairs <- raw_pairs %>% mutate(gene_name = str_to_upper(gene_name))
# # map_pairs <- map_pairs %>% mutate(gene_name = str_to_upper(gene_name))
# 
# # ---- 4) (Optional) Deduplicate pairs within each table ----
# raw_pairs_distinct <- distinct(raw_pairs)
# map_pairs_distinct <- distinct(map_pairs)
# 
# # ---- 5) Intersection: pairs present in both ----
# common_pairs <- inner_join(
#   raw_pairs_distinct,
#   map_pairs_distinct,
#   by = c("gene_id", "gene_name")
# )
# 
# # ---- 6) Pairs unique to each table (useful diagnostics) ----
# only_in_raw <- anti_join(raw_pairs_distinct, map_pairs_distinct,
#                          by = c("gene_id", "gene_name"))
# only_in_map <- anti_join(map_pairs_distinct, raw_pairs_distinct,
#                          by = c("gene_id", "gene_name"))
# 
# # ---- 7) Quick counts ----
# n_common    <- nrow(common_pairs)
# n_only_raw  <- nrow(only_in_raw)
# n_only_map  <- nrow(only_in_map)
# 
# list(
#   common_pairs = common_pairs,
#   only_in_raw  = only_in_raw,
#   only_in_map  = only_in_map,
#   summary = tibble(
#     common = n_common,
#     only_in_raw = n_only_raw,
#     only_in_map = n_only_map
#   )
# )


# # Set actual column names as above
# id_raw   <- raw_counts[["GeneID"]]
# name_raw <- raw_counts[["Gene name"]]
# id_map   <- counts_mapped[["ensembl_gene_id"]]
# name_map <- counts_mapped[["hgnc_symbol"]]
# 
# normalize <- function(x) trimws(gsub("\\s+", " ", as.character(x)))
# 
# key_raw <- paste(normalize(id_raw), normalize(name_raw), sep = "||")
# key_map <- paste(normalize(id_map), normalize(name_map), sep = "||")
# 
# common_keys <- intersect(unique(key_raw), unique(key_map))
# 
# # Turn back into a data.frame of pairs
# if (length(common_keys)) {
#   split_keys <- strsplit(common_keys, "\\|\\|", fixed = FALSE)
#   common_pairs_base <- data.frame(
#     gene_id   = vapply(split_keys, `[`, character(1), 1),
#     gene_name = vapply(split_keys, `[`, character(1), 2),
#     stringsAsFactors = FALSE
#   )
# } else {
#   common_pairs_base <- data.frame(gene_id = character(0), gene_name = character(0))
# }
# 
# common_pairs_base


# library(dplyr)
# 
# # Extract and rename columns to match
# df1 <- raw_counts %>%
#   dplyr::select(id1 = `GeneID`, name1 = `Gene name`) %>%
#   distinct()
# 
# df2 <- counts_mapped %>%
#   dplyr::select(id2 = ensembl_gene_id, name2 = hgnc_symbol) %>%
#   distinct()
# 
# # This will return only the pairs present in both
# matching_pairs <- inner_join(df1, df2, by = c("id1" = "id2", "name1" = "name2"))
# 
# # To see how many matches you found:
# nrow(matching_pairs)
```

```{r}
# # save intermmediate objects
# saveRDS(list(counts = counts, 
#              meta = meta, 
#              mappings = counts_mapped[,1:2]), 
#         file = "data/d1_counts_and_meta.rds")
```

## 4 Differential Expression Analysis

### 4.1 DE Analysis with EdgeR

### 4.2 Visualizing Differentially Expressed Genes (DEGs)
- MA
- Volcano
- Heatmap

```{r}
# obj <- readRDS("data/d1_qc_objects.rds")
# dge_f <- obj$dge_f
# meta <- obj$meta
# mappings <- obj$mappings
# 
# design <- obj$design
# 
# # Contrasts
# contr <- makeContrasts(
#   `3CTG_vs_NT` = treatment3CTG - treatmentNT,
#   `20CTG_vs_NT` = treatment20CTG - treatmentNT,
#   `20CTG_vs_3CTG` = treatment20CTG - treatment3CTG,
#   levels = design
# )
# 
# # Estimate dispersion and fit QL model
# edge <- edgeR::estimateDisp(dge_f, design)
# fit <- edgeR::glmQLFit(edge, design)
# 
# plotBCV(edge)
# 
# qlf_3_vs_nt <- edgeR::glmQLFTest(fit, contrast = contr[, "3CTG_vs_NT"])
# qlf_20_vs_nt <- edgeR::glmQLFTest(fit, contrast = contr[, "20CTG_vs_NT"])
# 
# edgeR::topTags(qlf_20_vs_nt)
# 
# qlf_20_vs_3 <- edgeR::glmQLFTest(fit, contrast = contr[, "20CTG_vs_3CTG"])
# 
# edgeR::topTags(qlf_20_vs_3)
# 
# tab_3_vs_nt <- edgeR::topTags(qlf_3_vs_nt, n = Inf)$table %>% rownames_to_column("ensembl")
# tab_20_vs_nt <- edgeR::topTags(qlf_20_vs_nt, n = Inf)$table %>% rownames_to_column("ensembl")
# tab_20_vs_3 <- edgeR::topTags(qlf_20_vs_3, n = Inf)$table %>% rownames_to_column("ensembl")
# 
# head(tab_20_vs_nt)
# 
# volcano_df <- tab_20_vs_nt %>%
#   mutate(sig = FDR < 0.05)
# 
# ggplot(volcano_df, aes(x = logFC, y = -log10(PValue), color = sig)) +
#   geom_point(alpha = 0.6) +
#   theme_minimal() +
#   labs(
#     title = "EdgeR: 20CTG vs NT",
#     x = "log2 fold-change",
#     y = "-log10(p-value)"
#   )
# 
# tab <- topTags(qlf_20_vs_nt, n = Inf)$table %>%
#   rownames_to_column("gene") %>%
#   mutate(
#     neglog10P = -log10(PValue),
#     direction = case_when(
#       FDR < 0.05 & logFC > 0 ~ "Up",
#       FDR < 0.05 & logFC < 0 ~ "Down",
#       TRUE ~ "Not sig"
#     )
#   )
# 
# ggplot(tab, aes(x = logFC, y = neglog10P)) +
#   geom_point(aes(color = direction), alpha = 0.7, size = 1.2) +
#   scale_color_manual(values = c(
#     "Up" = "red",
#     "Down" = "blue",
#     "Not sig" = "grey70"
#   )) +
#   geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "steelblue") +
#   geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "steelblue") +
#   theme_minimal() +
#   labs(
#     title = "edgeR volcano: 20CTG vs NT",
#     x = "log2 fold change (logFC)",
#     y = "-log10(p-value)",
#     color = NULL
#   )
# 
# # Keep only the columns we need and ensure they are character
# map_df <- mappings %>%
#   transmute(
#     ensembl = as.character(ensembl_gene_id),
#     hgnc_symbol = as.character(hgnc_symbol)
#   ) %>%
#   distinct(ensembl, .keep_all = TRUE)
# 
# # Join mapping onto the edgeR results table (tab_20_vs_nt)
# tab_20_vs_nt_annot <- tab_20_vs_nt %>%
#   mutate(ensembl = as.character(ensembl)) %>%
#   left_join(map_df, by = "ensembl") %>%
#   mutate(
#     # Use HGNC when available, otherwise fall back to Ensembl
#     display_name = ifelse(!is.na(hgnc_symbol) & hgnc_symbol != "",
#                           hgnc_symbol,
#                           ensembl)
#   )
# 
# # Quick check: how many mapped?
# mean(!is.na(tab_20_vs_nt_annot$hgnc_symbol))
# 
# head(tab_20_vs_nt_annot[, c("ensembl","hgnc_symbol","display_name","logFC","FDR")])
# 
# top_n <- 50  # change to 25/100 as desired
# 
# top_hits_tbl <- tab_20_vs_nt_annot %>%
#   arrange(FDR, PValue) %>%
#   slice_head(n = top_n)
# 
# # Ensembl IDs for subsetting the matrix
# top_hits_ensembl <- top_hits_tbl$ensembl
# 
# # Row labels we want to show on the heatmap
# top_hits_labels <- top_hits_tbl$display_name
# 
# head(top_hits_tbl[, c("ensembl","display_name","logFC","FDR")])
# 
# # logCPM from counts for visualization
# # logCPM for visualization
# logcpm <- edgeR::cpm(counts, log = TRUE, prior.count = 1)
# 
# # Keep only genes present in the expression matrix
# keep <- top_hits_ensembl %in% rownames(logcpm)
# 
# hm_mat <- logcpm[top_hits_ensembl[keep], , drop = FALSE]
# 
# # Create labels aligned to hm_mat rows
# hm_labels <- top_hits_labels[keep]
# 
# # IMPORTANT: avoid duplicate rownames (HGNC symbols can repeat)
# # Make labels unique while preserving readability
# rownames(hm_mat) <- make.unique(hm_labels)
# 
# # Z-score per gene across samples for pattern visibility
# hm_z <- t(scale(t(hm_mat)))
# 
# dim(hm_z)
# 
# meta$cellline <- as.factor(meta$cellline)
# meta$treatment <- as.factor(meta$treatment)
# 
# # Set3 palettes (qualitative)
# treat_levels <- levels(meta$treatment)
# cell_levels  <- levels(meta$cellline)
# 
# treat_cols <- setNames(
#   brewer.pal(max(3, length(treat_levels)), "Set3")[seq_along(treat_levels)],
#   treat_levels
# )
# 
# cell_cols <- setNames(
#   brewer.pal(max(3, length(cell_levels)), "Set2")[seq_along(cell_levels)],
#   cell_levels
# )
# 
# ha <- HeatmapAnnotation(
#   treatment = meta$treatment,
#   cellline  = meta$cellline,
#   col = list(
#     treatment = treat_cols,
#     cellline  = cell_cols
#   ),
#   annotation_legend_param = list(
#     treatment = list(title = "Treatment"),
#     cellline  = list(title = "Cell line")
#   )
# )
# 
# col_fun <- circlize::colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))
# 
# Heatmap(
#   hm_z,
#   col = col_fun,
#   top_annotation = ha,
#   show_row_names = FALSE,
#   show_column_names = FALSE,
#   cluster_rows = TRUE,
#   cluster_columns = TRUE,
#   row_title = paste0("Top ", nrow(hm_z), " genes (edgeR: 20CTG vs NT)"),
#   column_title = "Samples"
# )
# 
# Heatmap(
#   hm_z[1:25,],
#   col = col_fun,
#   top_annotation = ha,
#   show_row_names = TRUE,         # <-- now shows HGNC symbols
#   row_names_gp = grid::gpar(fontsize = 8),
#   show_column_names = FALSE,
#   cluster_rows = TRUE,
#   cluster_columns = TRUE,
#   row_title = paste0("Top ", nrow(hm_z), " genes (edgeR: 20CTG vs NT)"),
#   column_title = "Samples"
# )
# 
# saveRDS(
#   list(
#     meta = meta,
#     counts_mapped_filtered = counts_mapped_filtered,
#     fit = fit,
#     qlf_3_vs_nt = qlf_3_vs_nt,
#     qlf_20_vs_nt = qlf_20_vs_nt,
#     qlf_20_vs_3 = qlf_20_vs_3,
#     tab_3_vs_nt = tab_3_vs_nt,
#     tab_20_vs_nt = tab_20_vs_nt,
#     tab_20_vs_3 = tab_20_vs_3
#   ),
#   file = "data/d1_edger_results.rds"
# )
```

## 5 Discussion

1.  [Why is the dataset of interest to you?](#q1)
2.  [What are the control and test conditions of the dataset?](#q2)
3.  [How many samples in each of the conditions of your dataset?](#q3)
4.  [Were there expression values that were not unique for specific genes? How did you handle these?](#q4)
5.  [Were there expression values that could not be mapped to current HUGO symbols?](#q5)
6.  [Were there any outliers in your dataset? How were they handled in the originating paper? How many outliers were removed?](#q6)
7.  [How did you handle replicates?](#q7)
8.  [What is the final coverage of your dataset?](#q8)

Additional "rough work"/trial code and reflections can be found [here](https://github.com/bcb420-2026/Christine_Cheng/wiki/Entry-5-%E2%80%90-Assignment-1).

## 6 References

```{r, include=FALSE, echo=FALSE}
# TODO: generate citation bib
print(citation("rmarkdown"), bibtex=TRUE)
print(citation("knitr"), bibtex=TRUE)
citation("BiocManager")
citation("GEOquery")
citation("dplyr")
print(citation("biomaRt"), bibtex=TRUE)

citation("SummarizedExperiment")  # TODO: if using data from getRNASeqData
citation("readr") # TODO: if reading raw counts file from getGEOSuppFiles
```
