---
title: "A1 - Data set selection, cleaning, normalization, and differential gene expression"
author: "Christine En-Tse Cheng"
date: "02-10-2026"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs/A1") })
bibliography: A1_files/A1_references.bib    # TODO: insert the citations/bibliography file
link-citations: true            # this is for the citations/bibliography 
nocite: '@*'
---
```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE, # suppressing warning messages
                      message = FALSE,
                      knitr.kable.max_rows = 6)

# install/attach packages
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

if (!require("GEOquery", quietly = TRUE))
  BiocManager::install("GEOquery")

if (!require("readr", quietly = TRUE))
  install.packages("readr")

if (!require("dplyr", quietly = TRUE))
  install.packages("dplyr")

if (!require("biomaRt", quietly = TRUE))
  install.packages("biomaRt")

library(dplyr)
library(biomaRt)

# library(SummarizedExperiment) # TODO: if using data from getRNASeqData
library(readr) # TODO: if reading raw counts file from getGEOSuppFiles
```

## Data Exploration
### Downloading the data set
```{r}
GSE_ID <- "GSE166605"
```

```{r}
# dataset_se <- GEOquery::getRNASeqData(GSE_ID)
# dataset_se
# metadata(dataset_se)
# rowData(dataset_se)
```
```{r}
# # insert check for previously downloaded dataset in tempdir() or specified destdir
# dataset_gse <- GEOquery::getGEO(GEO = GSE_ID, 
#                                 destdir = "~/projects/Christine_Cheng/A1_files",
#                                 AnnotGPL = TRUE)
# dataset_gse
```

```{r data}
# access raw files from GEO
dataset_suppFiles <- GEOquery::getGEOSuppFiles(GSE_ID,
                                               makeDirectory = FALSE,
                                               baseDir = "~/projects/Christine_Cheng/A1_files")

counts_file <- file.path("~/projects/Christine_Cheng/A1_files",
                        "GSE166605_D1-5_Neg_C_miR_155_RAW_counts.txt.gz")
stopifnot(file.exists(counts_file))

# read the data files

raw_counts <- readr::read_tsv(counts_file, show_col_types = FALSE)

names(raw_counts)[1:min(12, ncol(raw_counts))]
```

```{r}
counts <- raw_counts %>%
  dplyr::select(starts_with("D")) %>%
  as.data.frame()

rownames(counts) <- raw_counts$GeneID

counts <- as.matrix(counts)
storage.mode(counts) <- "integer"

```
Parsing count data to make metadata for design matrix
```{r metadata}
samples <- colnames(counts)[1:ncol(counts)]
cell_line  <- unlist(lapply(samples,FUN =
                             function(x){unlist(strsplit(x,split = " "))[2]}))       
treatment <- unlist(lapply(samples,FUN = 
                             function(x){unlist(strsplit(x,split = " "))[3]}))     
meta_df <- data.frame(samples, cell_line, treatment)
```

### Exploratory Statistics

```{r}
dim(counts)
```

### Data Trimming/Cleaning
```{r}
# # QC, filtering and identifier mapping
# # This chapter performs shared QC steps used by both limma-voom and edgeR and DEseq2.
# 
# obj <- readRDS("data/d1_counts_and_meta.rds")
# counts <- obj$counts
# meta <- obj$meta
# mappings <- obj$mappings
# 
# # Create a DGEList (edgeR data container)
# dge <- edgeR::DGEList(counts = counts)
# 
# # Attach metadata
# # (not required, but helpful)
# dge$samples <- cbind(dge$samples, meta)
# 
# dge$samples
# 
# # filter lowly expressed genes
# design <- model.matrix(~ 0 +  treatment+ cellline, data = meta)
# 
# keep <- edgeR::filterByExpr(dge, design)
# dge_f <- dge[keep, , keep.lib.sizes = FALSE]
# 
# dim(dge)
```


## Data Normalization
```{r}
# # TMM normalization
# dge_f <- edgeR::calcNormFactors(dge_f, method = "TMM")
# dge_f$samples
```

```{r}
# # exploratory MDS plot
# current_colors <- RColorBrewer::brewer.pal(n = length(unique(meta$treatment)), "Set2")
# names(current_colors) <- levels(factor(meta$treatment))
# 
# plotMDS(dge_f,
#   labels = meta$treatment,
#   col = current_colors[meta$treatment],
#   main = "MDS plot (colored by treatment)"
# )
# legend("topright",
#   legend = names(current_colors),
#   col = current_colors,
#   pch = 16,
#   bty = "n"
# )
# 
# # " " " - coloured by cell line
# current_colors_cellline <- RColorBrewer::brewer.pal(n = length(unique(meta$cellline)), "Set1")
# names(current_colors_cellline) <- levels(factor(meta$cellline))
# 
# plotMDS(dge_f,
#   labels = meta$cellline,
#   col = current_colors_cellline[meta$cellline],
#   main = "MDS plot (colored by cellline)"
# )
# legend("topright",
#   legend = names(current_colors_cellline),
#   col = current_colors_cellline,
#   pch = 16,
#   bty = "n"
# )
```

```{r}
# # log-CPM for visualization
# lcpm <- edgeR::cpm(dge_f, log = TRUE, prior.count = 1)
# summary(as.vector(lcpm))
```

```{r}
# saveRDS(list(dge_f = dge_f, 
#              meta = meta, 
#              design = design, 
#              lcpm = lcpm, 
#              mappings = mappings),
#         file = "data/d1_qc_objects.rds")
```

### Gene ID mapping

```{r id_mapping}
# strip_ensembl_version <- function(x) sub("\\..*$", "", x)
# 
# kable_head <- function(x, n = 5, caption = NULL) {
#   knitr::kable(utils::head(x, n), caption = caption)
# }
# 
# mapped_counts <- data.frame(ensembl_gene_id = rownames(counts),counts)
# ids<- mapped_counts$ensembl_gene_id
# 
# if (any(grepl("^ENSG", strip_ensembl_version(ids)))) {
#   library(biomaRt)
#   ensembl_ids <- unique(strip_ensembl_version(ids))
#   ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
# 
#   map <- getBM(
#     attributes = c("ensembl_gene_id", "hgnc_symbol"),
#     filters = "ensembl_gene_id",
#     values = ensembl_ids,
#     mart = ensembl
#   )
# 
#   kable_head(map, 10, paste(": mapping preview"))
# 
# counts_mapped <- mapped_counts %>%
#     left_join(map, by = "ensembl_gene_id") %>%
#   dplyr::select( ensembl_gene_id,hgnc_symbol, everything())
# 
# 
#   kable_head(counts_mapped[, 1:min(8, ncol(counts_mapped))], 5, paste(": mapped table preview"))
# }
```

```{r}
# # save intermmediate objects
# saveRDS(list(counts = counts, 
#              meta = meta, 
#              mappings = counts_mapped[,1:2]), 
#         file = "data/d1_counts_and_meta.rds")
```

## Differential Expression Analysis

## Discussion
1. Why is the dataset of interest to you?
2. What are the control and test conditions of the dataset?
3. How many samples in each of the conditions of your dataset?
4. Were there expression values that were not unique for specific genes? How did you handle these?
5. Were there expression values that could not be mapped to current HUGO symbols?
6. Were there any outliers in your dataset? How were they handled in the originating paper? How 7. many outliers were removed?
8. How did you handle replicates?
9. What is the final coverage of your dataset?

## References

```{r, include=FALSE, echo=FALSE}
# TODO: generate citation bib
print(citation("rmarkdown"), bibtex=TRUE)
print(citation("knitr"), bibtex=TRUE)
citation("BiocManager")
citation("GEOquery")
citation("dplyr")
print(citation("biomaRt"), bibtex=TRUE)

citation("SummarizedExperiment")  # TODO: if using data from getRNASeqData
citation("readr") # TODO: if reading raw counts file from getGEOSuppFiles
```
