---
title: "A2 - Dataset Pathway and Network Analysis"
author: "Christine En-Tse Cheng"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
params:
  working_dir: ./A2_files
  # gsea_jar: ../../GSEA_4.4.0/gsea-cli.sh  # TODO: switch to docker ver.
  gsea_jar: "C:/Users/chris/Downloads/BCB420H1-S/GSEA/GSEA_4.4.0/gsea-cli.bat"
  run_gsea: True
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs/A2") })
bibliography: A2_files/A2_references.bib
link-citations: true            # this is for the citations/bibliography 
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      knitr.kable.max_rows = 6)

# helper function to present formatted data table previews (default: 6 rows)
kable_head_helper <- function() {
  i <- 0
  function(x, n = 6, caption = NULL) {
    i <<- i + 1
    full_caption <- paste0("**Table ", i, ".** ", caption)
    knitr::kable(utils::head(x, n), caption = full_caption)
  }
}
kable_head <- kable_head_helper()

working_dir <- params$working_dir
# data_dir <- working_dir
```

```{r packages}
# install/attach packages
# if (!require("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")

if (!require("gprofiler2", quietly = TRUE)) {
  install.packages("gprofiler2")
  library(gprofiler2)
}

if (!require("GSA", quietly = TRUE)) {
  install.packages("GSA")
  library(GSA)
}

if (!require("RCurl", quietly = TRUE)) {
  install.packages("RCurl")
  library(RCurl)
}  
```
```{r}
# setwd("C:/Users/chris/Downloads/BCB420H1-S")
```


## 1 Introduction

TODO: summary of the normalization and scoring done in the first assignment. Assume that the person reading the report has not read your assignment #1 report. Including basic statistics from that analysis will be helpful. (for example, data downloaded from GEO with id X, …).
  - TODO: actually do the normalization properly...I ran the TMM thing and got graphs to show what normalized data would look like, but I don't think I actually ended up using the normalized data -_-

You can include your previous notebooks as subsections of this report by following the instructions outlined here: <https://yihui.org/knitr/demo/child/> This report describes the selection, data preparation, normalization, and differential expression analysis of a publicly available human gene expression dataset. The workflow follows best practices for RNA-seq analysis as outlined by (@dawadi_bench_2025).

This report performs pathway and network analysis on the differentially expressed genes (DEGs) reported by [Assignment #1](https://bcb420-2026.github.io/Christine_Cheng/A1/A1_ChristineCheng.html).

```{r dge_results}
# dge_genes <- readRDS("./dge_results.rds")
a1_results <- readRDS("./A1_outputs.rds")
# full_genelist <- a1_results$id_mappings
dge_results <- a1_results$dge_results
kable_head(dge_results, caption = "Preview of top DE genes")
```

Below is a quick overview of the DEGs reported previously:

```{r dge_stats}
# upregulated DEGs
upDEGs <- which(dge_results$FDR < 0.05 & dge_results$logFC > 0)
cat("# of upregulated differentially expressed genes: ", length(upDEGs))

# downregulated DEGs
downDEGs <- which(dge_results$FDR < 0.05 & dge_results$logFC < 0)
cat("# of downregulated differentially expressed genes: ", length(downDEGs))
```

## 2 Thresholded over-representation analysis (ORA) / Gene List Enrichment Analysis {#q2.1}

First, thresholded over-representation analysis was performed with the `gprofiler2` package. I chose this method because ...

```{r gprofiler_prep}
# genelist_file <- dge_results
max_gs_size <- 250  # TODO: varying, maybe?? and justification of value
min_gs_size <- 3 # TODO: justification
min_intersection <- 3 # TODO: justification

orgnsm_binom <- "hsapiens"  # organism parameter for g:Profiler

# checking for local version of gmt file, before accessing and downloading from Baderlab
dest_gmt_file = "https://download.baderlab.org/EM_Genesets/September_01_2025/Human/symbol/Human_GOBP_AllPathways_noPFOCR_no_GO_iea_September_01_2025_symbol.gmt"
local_gmt <- file.path(working_dir,
                       "Human_GOBP_AllPathways_noPFOCR_no_GO_iea_September_01_2025_symbol.gmt")
if (!file.exists(local_gmt)) {
  download.file(dest_gmt_file, destfile = local_gmt, mode = "wb")
}

# access and download g:Profiler .gmt file
gprofiler_gmt_url <- "https://biit.cs.ut.ee/gprofiler/static/gprofiler_full_hsapiens.name.gmt"
gprofiler_ver <- get_version_info(organism = orgnsm_binom)

gprofiler_gmt_filename <- file.path(working_dir,
                                    paste("gprofiler_full", orgnsm_binom,
                                          gprofiler_ver$gprofiler_version, sep="_",
                                          ".name.gmt"))

if (!file.exists(gprofiler_gmt_filename)) {
  download.file(url = gprofiler_gmt_url,
                destfile = gprofiler_gmt_filename)
  }
```

```{r run_gprofiler}
# TODO: build query genelist: should be thresholded (i.e., only a subset of genes)
gene_list <- dge_results[dge_results$FDR < 0.5, ]$hgnc_symbol

# g:Profiler (default geneset) query
gprofiler_results <- gprofiler2::gost(query = gene_list,
                                      significant = FALSE, # to return all results; not just significance thresholded results
                                      ordered_query = FALSE, # TODO: check whether query gene list is ordered by significance or not
                                      exclude_iea = TRUE,
                                      correction_method = "fdr",
                                      organism = orgnsm_binom,
                                      sources = c("REAC", "WP", "GO:BP"))
enrichment_results <- gprofiler_results$result
```

```{r}
# TODO: Run the analysis using the up-regulated set of genes, and the down-regulated set of genes separately
```

### 2.1 ORA method

### 2.2 Annotation dataset

### 2.3 Discussion

## 3 Non-thresholded Gene Set Enrichment Analysis (GSEA) {#q3.1}

Next, non-thresholded gene set enrichment analysis using GSEA from R was conducted on the same set of DEGs from assignment #1. I chose this method because ...

```{r gsea_setup}
#path to GSEA jar 
gsea_jar <- params$gsea_jar

#directory where all the data files are found and should be written to
working_dir <- params$working_dir
output_dir <- params$working_dir

#The name to give the analysis in GSEA
analysis_name <- "miR155_vs_NegCtrl"

run_gsea <- params$run_gsea
```

```{r gmt_file}
gmt_url <- "https://download.baderlab.org/EM_Genesets/current_release/Human/symbol/"

filenames <- getURL(gmt_url)
tc <- textConnection(filenames)
contents <- readLines(tc)
close(tc)

rx <- gregexpr("(?<=<a href=\")(.*.GOBP_AllPathways_noPFOCR_no_GO_iea.*.)(.gmt)(?=\">)",
              contents, perl = TRUE)
gmt_filename <- unlist(regmatches(contents, rx))
gmt_file <- file.path(working_dir, gmt_filename)

if (!file.exists(gmt_file)) {
  download.file(paste(gmt_url, gmt_filename, sep = ""), destfile = gmt_file, mode = "wb")
}

# # set the gmt file to Bader Lab's no OCR/IEA gmt file from September 01, 2025:
# dest_gmt_file = "https://download.baderlab.org/EM_Genesets/September_01_2025/Human/symbol/Human_GOBP_AllPathways_noPFOCR_no_GO_iea_September_01_2025_symbol.gmt"
# 
# local_gmt <- file.path(working_dir, "Human_GOBP_AllPathways_noPFOCR_no_GO_iea_September_01_2025_symbol.gmt")
# 
# if (!file.exists(local_gmt)) {
#   download.file(dest_gmt_file, destfile = local_gmt, mode = "wb")
# }
```

```{r rank_genes}
# rank file to use in GSEA analysis.  
gene_ranks = sign(dge_results$logFC) * -log10(dge_results$PValue)

#create ranks file
ranked_genelist <- cbind(dge_results$hgnc_symbol, gene_ranks)
colnames(ranked_genelist) <- c("GeneName", "Rank")

#sort ranks in decreasing order
ranked_genelist <- ranked_genelist[order(as.numeric(ranked_genelist[, 2]), decreasing = TRUE), ]

# write to file as a table
rnk_path <- file.path(working_dir, paste0(analysis_name, "_Ranked_GeneList.rnk"))
write.table(ranked_genelist, file = rnk_path, sep = "\t",
            col.name = TRUE, row.names = FALSE, quote = FALSE)
```

```{r run_gsea}
if(run_gsea){
  command <- paste("", gsea_jar,  
                   "GSEAPreRanked -gmx", gmt_file, 
                   "-rnk" , rnk_path, 
                   "-collapse false -nperm 1000 -scoring_scheme weighted", 
                   "-rpt_label ", analysis_name,
                   "  -plot_top_x 20 -rnd_seed 12345  -set_max 200",  
                   " -set_min 15 -zip_report false ",
                   " -out" , output_dir, 
                   " > gsea_output.txt", sep= " ")
  # TODO: uncomment for docker version
  # system(command)
  shell(command)
}
```

### 3.1 GSEA method and genesets

### 3.2 Discussion

## 4 Cytoscape visualization

### 4.1 Enrichment Map

### 4.2 Network Annotation

### 4.3 Theme Network

### 4.4 Discussion

TODO: link journal entry Additional "rough work"/trial code and reflections can be found [here](https://github.com/bcb420-2026/Christine_Cheng/wiki/Assignment-2).

## 5 Question Collections

### 5.1 Thresholded over-representation analysis (ORA)

1.  [Which method did you choose and why?](#q2.1)

2.  [What annotation data did you use and why? What version of the annotation are you using?](#q2.2)

3.  [How many genesets were returned with what thresholds?](#q2.3)

4.  [How do the analysis results on the separated up-regulated and down-regulated gene lists compare to using the whole list of differentially expressed genes together?](#q2.4)

5.  [Do the over-representation results support conclusions or mechanism discussed in the original paper?](#q2.5)

6.  [Can you find evidence, i.e. publications, to support some of the results that you see. How does this evidence support your results.](#q2.6)

### 5.2 Non-thresholded Gene Set Enrichment Analysis (GSEA)

1.  [What method did you use? What genesets did you use?](#q3.1)

2.  [Summarize your enrichment results.](#q3.2)

3.  [How do these results compare to the results from the thresholded analysis? Compare qualitatively. Is this a straight forward comparison? Why or why not?](#q3.3)

### 5.3 Cytoscape visualization

1.  [How many nodes and how many edges were in the resulting enrichment map? What thresholds were used to create this map?](#q4.1)

2.  [What parameters did you use to annotate the network?](#q4.2)

3.  [Make a publication ready figure.](#q4.3)

4.  [What are the major themes present in the collapse to a theme network? Do they fit with the model? Are there any novel pathways or themes?](#q4.4)

5.  [Do the enrichment results support conclusions or mechanism discussed in the original paper? How do these results differ from the results you got from thresholded methods?](#q4.5)

6.  [Can you find evidence, i.e. publications, to support some of the results that you see? How does this evidence support your result?](#q4.6)

## 6 Discussion and Conclusion

## 7 References

```{r}
# citation("gprofiler2")
# citation("GSA")
# citation("RCurl")
```
